---
title: "Data & Setup"
bibliography: course_refs.bib  
number-sections: false
---

<!-- 
Note for Training Developers:
We provide instructions for commonly-used software as commented sections below.
Uncomment the sections relevant for your materials, and add additional instructions where needed (e.g. specific packages used).
Note that we use tabsets to provide instructions for all three major operating systems.
-->

::: {.callout-tip level=2}
## Workshop Attendees

If you are attending one of our workshops, we will provide a training
environment with all of the required software and data. If you want to setup
your own computer to run the analysis demonstrated on this course, you can
follow the instructions below.
:::

## Data

The data used in these materials is provided as a zip file. 
Download and unzip the folder to your Desktop to follow along with the materials.

<!-- Note for Training Developers: add the link to 'href' -->
<a href="">
  <button class="btn"><i class="fa fa-download"></i> Download</button>
</a>

## Working on your own laptop

If you wish to work on your own laptop during the course and not the training
machines you will need to have the latest stable versions of R, RStudio and
Bioconductor including the packages listed below.

### R and RStudio

This lesson assumes that you have current versions of the following installed on 
your computer:

1. The [R software](https://cran.r-project.org/mirrors.html) itself, and
2. [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/#download).

R and RStudio require separate downloads and installations. R itself refers to 
the underlying language and computing environment required to write and read 
this language. RStudio is a graphical integrated development environment which
essentially makes running R much easier and more interactive.  

<!-- ### Quarto -->

<!-- To develop and render the course materials website, you will need to install _Quarto_: -->

<!-- - Download and install [Quarto](https://quarto.org/docs/get-started/) (available for all major OS).  -->
<!-- - If you are developing materials using executable `.qmd` documents, it is recommended that you also install the extensions for your favourite IDE (e.g. RStudio, VS Code). -->
<!-- - If you are developing materials using **JupyterLab** or **Jupyter Notebooks**, please install [Jupytext](https://jupytext.readthedocs.io/en/latest/install.html). -->
<!--   - Use the [paired notebook](https://jupytext.readthedocs.io/en/latest/paired-notebooks.html) feature to have synchronised `.ipynb`/`.qmd` files. Only `.qmd` files should be pushed to the repository (`.ipynb` files have been added to `.gitignore`). -->


::: {.tabset group="os"}

#### Windows

Download and install all these using default options:

- [R](https://cran.r-project.org/bin/windows/base/release.html)
- [RTools](https://cran.r-project.org/bin/windows/Rtools/)
- [RStudio](https://www.rstudio.com/products/rstudio/download/#download)

#### Mac OS

Download and install all these using default options:

- [R](https://cran.r-project.org/bin/macosx/)
- [RStudio](https://www.rstudio.com/products/rstudio/download/#download)

#### Linux

- Go to the [R installation](https://cran.r-project.org/bin/linux/) folder and look at the instructions for your distribution.
- Download the [RStudio](https://www.rstudio.com/products/rstudio/download/#download) installer for your distribution and install it using your package manager.

:::

### R package installation

In this workshop we make use of open-source software from the 
[R](https://www.r-project.org) [Bioconductor](https://bioconductor.org) 
[@Huber2015] project. The [Bioconductor initiative](https://bioconductor.org) 
provides [R](https://www.r-project.org) software packages dedicated
to the processing of high-throughput complex biological data. Packages are
open-source, well-documented and benefit from an active community of developers.

Detailed instructions for the installation of Bioconductor packages are
documented on the [Bioconductor Installation page](http://bioconductor.org/install/). 
The main packages required for this workshop are installed using the code below.
Additional packages required for downstream statistics and interpretation are
installed as required.

```{r install_packages, eval = FALSE, message = FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("QFeatures",
                       "ggplot2",
                       "stringr"
                       "NormalyzerDE",
                       "corrplot",
                       "Biostrings",
                       "limma",
                       "impute",
                       "dplyr",
                       "tibble",
                       "org.Hs.eg.db",
                       "clusterProfiler",
                       "enrichplot"))
```

After installation, each package must be loaded before it can be used in the R
session. This is achieved via the `library` function. Here we load all packages 
included in this course.

```{r load_packages, message = FALSE, results = "hide", message=FALSE, warning=FALSE}
library("QFeatures")
library("ggplot2")
library("stringr")
library("dplyr")
library("tibble")
library("NormalyzerDE")
library("corrplot")
library("Biostrings")
library("limma")
library("org.Hs.eg.db")
library("clusterProfiler")
library("enrichplot")
```


<!--
### Conda

Open a terminal and run:

```bash
wget -q -O - https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh
bash Miniconda3-latest-Linux-x86_64.sh -b
rm Miniconda3-latest-Linux-x86_64.sh
conda init
conda config --add channels defaults; conda config --add channels bioconda; conda config --add channels conda-forge; conda config --set channel_priority strict
conda install -y mamba
```

Note: Windows users can use WSL2 (see @wsl).
-->

<!--
### Singularity

::: {.panel-tabset group="os"}

#### Windows

You can use _Singularity_ from the _Windows Subsystem for Linux_ (see @wsl).  
Once you setup WSL, you can follow the instructions for Linux.

#### Mac OS

Singularity is [not available for Mac OS](https://docs.sylabs.io/guides/3.0/user-guide/installation.html#install-on-windows-or-mac).

#### Linux

These instructions are for _Ubuntu_ or _Debian_-based distributions[^1].

[^1]: See the [Singularity documentation page](https://docs.sylabs.io/guides/3.0/user-guide/installation.html#install-on-linux) for other distributions.

```bash
sudo apt update && sudo apt upgrade && sudo apt install runc
CODENAME=$(lsb_release -c | sed 's/Codename:\t//')
wget -O singularity.deb https://github.com/sylabs/singularity/releases/download/v3.10.2/singularity-ce_3.10.2-${CODENAME}_amd64.deb
sudo dpkg -i singularity.deb
rm singularity.deb
```

:::
-->


<!-- 
### Visual Studio Code

::: {.panel-tabset group="os"}

#### Windows

- Go to the [Visual Studio Code download page](https://code.visualstudio.com/Download) and download the installer for your operating system. 
  Double-click the downloaded file to install the software, accepting all the default options. 
- After completing the installation, go to your Windows Menu, search for "Visual Studio Code" and launch the application. 
- Go to "_File > Preferences > Settings_", then select "_Text Editor > Files_" on the drop-down menu on the left. Scroll down to the section named "_EOL_" and choose "_\\n_" (this will ensure that the files you edit on Windows are compatible with the Linux operating system).

#### Mac OS

- Go to the [Visual Studio Code download page](https://code.visualstudio.com/Download) and download the installer for Mac.
- Go to the Downloads folder and double-click the file you just downloaded to extract the application. Drag-and-drop the "Visual Studio Code" file to your "Applications" folder. 
- You can now open the installed application to check that it was installed successfully (the first time you launch the application you will get a warning that this is an application downloaded from the internet - you can go ahead and click "Open").

#### Linux (Ubuntu)

- Go to the [Visual Studio Code download page](https://code.visualstudio.com/Download) and download the installer for your Linux distribution. Install the package using your system's installer.

:::
 -->

# References
