---
title: Obtaining Reactome pathway annotations
bibliography: ../../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Understand how to retrieve Reactome pathway annotations for a set of UniProt protein IDs
* Appreciate why Reactome annotations must be expanded to include all ancestor (parent) pathway terms
* Be able to map Reactome pathway IDs to human-readable pathway names using `reactome.db`

:::

::: {.callout-note}
This notebook describes how to obtain and process the Reactome pathway
annotations used in the [Functional enrichment analysis](../08_functional_enrichment.html)
section. The pre-computed objects (`reactome_annot_list` and `reactome_map`)
are loaded directly in that lesson, so this material is not required to
follow the main course. It is provided here for completeness and
reproducibility.
:::

```{r, eval=TRUE, include=FALSE}
library("here")
library("UniProt.ws")
library("reactome.db")
library("tidyr")
library("dplyr")
```

## Retrieving Reactome annotations from UniProt

We want to retrieve Reactome pathway annotations for all proteins identified in
our dataset. The starting point is the set of UniProt accession IDs stored in
the `dia_qf` object we created in earlier lessons.

```{r}
load(here("course_files/dia/preprocessed/lesson04.rda"), verbose = TRUE)
```

We extract the unique protein accession IDs from the row data of the
normalised protein assay. These are the identifiers we will use to query
UniProt for Reactome pathway annotations.

```{r}
all_proteins <- rownames(dia_qf[["norm_proteins"]])
```

We use the `mapUniProt` function from the `UniProt.ws` package to query
the UniProt REST API. We request three columns: the UniProt accession
(`accession`), the primary gene name (`gene_primary`), and any associated
Reactome pathway identifiers (`xref_reactome`). The query may take a minute
or two depending on the number of proteins and network speed.

```{r}
reactome_annot <- mapUniProt(
  from    = "UniProtKB_AC-ID",
  to      = "UniProtKB",
  columns = c("accession", "gene_primary", "xref_reactome"),
  query   = all_proteins[all_proteins != ""]
)
```

We can check whether any of our input proteins were not returned by the
query, which can happen if an accession has been retired or merged in
UniProt.

```{r}
setdiff(all_proteins, reactome_annot$Entry)
```

The `Reactome` column returned by `mapUniProt` may contain multiple
semicolon-separated pathway IDs for each protein. We use `separate_rows`
to convert these into one row per protein–pathway pair, then remove any
rows where the Reactome ID is empty.

```{r}
reactome_annot <- reactome_annot %>%
  separate_rows(Reactome, sep = ";\\s*") %>%
  filter(Reactome != "")
```

In some cases the Reactome ID contains additional text (e.g. a protein
isoform name appended after a space). We strip this to retain only the
accession-style ID (format: `R-HSA-XXXXXXX`).

```{r}
reactome_annot <- reactome_annot %>%
  mutate(Reactome = sub(" .*", "", Reactome))
```

## Expanding annotations to include ancestor pathways

Reactome is a hierarchical database: pathways are nested within broader
pathways. For example, a specific signalling event will belong to a
signalling cascade, which in turn belongs to a broader biological process.
When performing gene set enrichment analysis, we want a protein to be
associated not only with its directly annotated pathway, but also with all
of its ancestor (parent) pathways in the hierarchy. Without this expansion,
we would miss enrichment signals at broader pathway levels.

We download the parent–child relationship table directly from Reactome. This
file lists every pathway pair where one pathway is a sub-pathway of another.

```{r}
hierarchy <- read.delim(
  "https://reactome.org/download/current/ReactomePathwaysRelation.txt",
  header    = FALSE,
  col.names = c("parent", "child")
)
```

We restructure this into a lookup table mapping each child pathway to its
direct parents.

```{r}
child_to_parents <- split(hierarchy$parent, hierarchy$child)
```

We then write a recursive function that, given a pathway ID, returns all
of its ancestors by repeatedly looking up parents until no further parents
are found. The function also includes the input pathway ID itself in the
result, so each pathway is associated with itself and all of its ancestors.

```{r}
get_reactome_ancestors <- function(pathway_id, lookup = child_to_parents) {
  parents <- lookup[[pathway_id]]
  if (is.null(parents)) return(character(0))

  all_ancestors <- parents
  for (parent in parents) {
    all_ancestors <- c(all_ancestors, get_reactome_ancestors(parent, lookup))
  }
  unique(c(pathway_id, all_ancestors))
}
```

We apply this function to all unique Reactome IDs in our annotation table,
producing a named list where each element contains the full set of pathway
IDs (direct and ancestral) for one input pathway.

```{r}
ancestor_list <- sapply(
  unique(reactome_annot$Reactome),
  get_reactome_ancestors
)

names(ancestor_list) <- unique(reactome_annot$Reactome)
```

We convert this list to a long-format data frame and merge it with the
original annotation table. The result, `reactome_annot_expanded`, has one
row per protein–pathway pair, where the pathways now include all ancestor
terms.

```{r}
ancestor_df <- data.frame(
  Reactome          = rep(names(ancestor_list), lengths(ancestor_list)),
  Reactome_expanded = unname(unlist(ancestor_list))
)

reactome_annot_expanded <- merge(reactome_annot, ancestor_df, by = "Reactome")
```

## Mapping Reactome IDs to pathway names

Reactome pathway IDs (e.g. `R-HSA-123456`) are not human-readable. We use
the `reactome.db` Bioconductor package to map these IDs to descriptive
pathway names. This package provides a direct offline lookup and does not
require any API calls.

We look up names for all unique expanded pathway IDs, then tidy up the
result by removing the `"Homo sapiens: "` prefix that Reactome prepends to
all human pathway names.

```{r}
reactome_ids <- unique(reactome_annot_expanded$Reactome_expanded)

reactome_map <- as.data.frame(reactomePATHID2NAME[reactome_ids]) %>%
  dplyr::rename(Reactome = DB_ID, Pathway = path_name) %>%
  mutate(Pathway = sub("Homo sapiens: ", "", Pathway))

head(reactome_map)
```

## Building the pathway–protein list

For GSEA with `fgsea`, we need the annotations in the form of a named list
where each element is a character vector of UniProt IDs belonging to that
pathway. We group the expanded annotation table by expanded Reactome ID and
extract the unique protein accessions for each pathway.

```{r}
reactome_annot_list <- reactome_annot_expanded %>%
  group_by(Reactome_expanded) %>%
  group_map(function(x, ...) {
    unique(x$Entry)
  })

names(reactome_annot_list) <- group_keys(
  group_by(reactome_annot_expanded, Reactome_expanded)
)$Reactome_expanded
```

## Saving the objects

We save both objects as `.rds` files so they can be loaded quickly in the
main functional enrichment notebook without repeating the API queries and
processing steps above.

```{r}
saveRDS(reactome_annot_list,
        here("course_files/dia/preprocessed/reactome_annot_list.rds"))
saveRDS(reactome_map,
        here("course_files/dia/preprocessed/reactome_map.rds"))
```

::: {.callout-tip}
#### Key Points

* `mapUniProt` from the `UniProt.ws` package retrieves Reactome pathway annotations for a set of UniProt accession IDs via the UniProt REST API
* Reactome annotations must be expanded to include all ancestor pathways using the parent–child hierarchy file from Reactome, so that enrichment can be detected at all levels of the pathway hierarchy
* `reactome.db` provides an offline mapping from Reactome pathway IDs to human-readable pathway names
* The resulting `reactome_annot_list` (pathway → proteins) and `reactome_map` (ID → name) objects are saved as `.rds` files for use in downstream GSEA

:::
