---
title: Import and infrastructure
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Understand the structure of `SummarizedExperiment` and `QFeatures` objects and how the two are related
* Understand how DIA-NN output data is structured and how it can be imported into R
* Know how to use `readQFeaturesFromDIANN` to create a `QFeatures` object from DIA-NN output
* Be able to access sample metadata via `colData` in a `QFeatures` object
* Be able to explore the data within a `QFeatures` object using `assay`, `rowData`, and `colData`

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("arrow")
library("readxl")
library("stringr")
library("QFeatures")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")
```


## Importing DIA-NN output data

The raw DIA mass spectrometry data was processed using DIA-NN, which outputs a
report file in `.parquet` format. This file contains precursor- and protein-level
identifications and quantities for each sample. We begin by reading this file
into R using the `read_parquet` function from the `arrow` package.

```{r}
diann_df <- read_parquet(here("course_files/dia/DIANNout/monkeypox_plasma_proteomes.parquet"))
```

### Cleaning run names

DIA-NN pastes together the name of the `.wiff2` folder and the `.wiff.scan`
files into the `Run` column. We need to clean up these run names so that they
match the `sample_id` field in our metadata.

```{r}
### Clean Run names
print(head(unique(diann_df$Run), 2))

# DIANN has pasted the name of the .wiff2 folder as well as the .wiff.scan files together.
# So we need to clean up the Run names so that they match the metadata `sample_id` field.
diann_df <- diann_df %>%
  mutate(Run = sub('\\..*', '', Run))

print(head(unique(diann_df$Run), 2))
```

{{< include ../_shared/_pipe.md >}}

## Importing sample metadata

We import the sample metadata from an Excel file. This provides information
about each sample, including which group (MPXV, COVID-19, or Control) each
run belongs to. We rename the `sample_id` column to `runCol` so it is
compatible with the `readQFeaturesFromDIANN` function.

```{r}
colData <- read_excel(here("course_files/dia/metadata/monkeypox_metadata.xlsx")) %>%
  dplyr::rename(runCol = sample_id)
```

{{< include ../_shared/_qfeatures_structure.md >}}

The best way to get our head around the `QFeatures` infrastructure is to import
our data into a `QFeatures` object and start exploring.

## Reading into a `QFeatures` object

We use the `readQFeaturesFromDIANN` function to import the cleaned DIA-NN data
into a `QFeatures` object. This function creates one `SummarizedExperiment`
per run, with each assay containing the precursor-level quantification data
for that run. The `quantCols` argument specifies which column contains the
quantitative data, and `fnames` specifies the column to use as feature names.

```{r}
dia_qf <- QFeatures::readQFeaturesFromDIANN(
  diann_df, quantCols = "Precursor.Quantity", fnames = "Precursor.Id",
  runCol = 'Run', colData = colData)


```



## Exploring the data

When we inspect our `QFeatures` object, we see that it contains 31 assays, each corresponding to a different sample run. Each assay contains the
precursor-level quantification data for that run, with feature names given by the `Precursor.Id` column from the original DIA-NN output.

```{r}
dia_qf

head(assay(dia_qf[[1]]))
```

::: {.callout-discussion}
#### Discussion

Why does `readQFeaturesFromDIANN` create one `SummarizedExperiment` per run,
rather than a single combined assay across all runs from the outset?

:::

::: {.callout-answer collapse="true"}

The DIA-NN output contains per-run quality metrics
stored in the `rowData` of each precursor. These are run-specific values and
would be ambiguous or impossible to retain if all runs were immediately merged
into a single matrix. By keeping one `SummarizedExperiment` per run, we can
inspect and filter on these per-run metrics before joining the data. We will
join the per-sample assays into a single combined precursor matrix in the next
lesson, once filtering is complete.

:::

### How many precursors per sample?

Each sample will have a different number of precursors identified and quantified. We can explore
this by inspecting the number of rows in each per-sample assay, which corresponds to the number of precursors quantified in that sample.

:::{.callout-exercise}
#### Challenge: Precursors per sample

{{< level 2 >}}

1. Find a function to determine how many precursors are quantified in each
sample. Use it to plot a histogram of precursor counts across samples.

2. Create a bar plot showing the number of precursors identified in each
sample, coloured by group. Based on your plot:

- Which group tends to have the most precursors identified?
- Is there any sample that looks like an outlier?

<details><summary>Hint</summary>Browse the `QFeatures` documentation with
`?QFeatures` to find a suitable function for obtaining per-sample feature
counts.</details>

::: {.callout-answer collapse="true"}

**Task 1**

```{r, fig.height=4, fig.width=4}
hist(nrows(dia_qf),
     xlab="Precursors",
     main="Precursor counts per sample")

```

**Task 2**

```{r}
data.frame(
  n_precursors = nrows(dia_qf),
  sample = names(nrows(dia_qf))) %>%
  merge(data.frame(colData(dia_qf)), by.x = "sample", by.y = "runCol") %>%
  ggplot(aes(x = shortname, y = n_precursors, fill = group)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Number of precursors") +
  coord_flip() +
  theme_classic()
```

:::

:::


### Exploring the row data

The `rowData` of each per-sample `SummarizedExperiment` contains quality metrics
from DIA-NN, including the Full-Width at Half Maximum (FWHM) of the
chromatographic peaks. We can extract this information across all runs using the
`rbindRowData` function. If we merge with the column data, we can then visualise
the distribution of FWHM values per sample, coloured by group.

```{r, fig.width=5, fig.height=5}
cdata <- data.frame(colData(dia_qf))

rdata <- rbindRowData(dia_qf, i = names(dia_qf)) %>%
  data.frame() %>%
  merge(cdata, by.x = "Run", by.y = "runCol")

rdata %>%
  ggplot(aes(FWHM, group = Run, colour = group)) +
  geom_density() +
  theme_classic()
```

:::{.callout-exercise}
#### Challenge: Exploring rowData quality metrics

{{< level 2 >}}

The `rdata` object contains many columns from the DIA-NN `rowData` beyond FWHM.

1. Use `str(rdata)` to identify other numerical quality metric columns.
2. Choose one metric and create a plot to visualise its distribution across
   samples, coloured by group.
3. Do any metrics appear to differ systematically between groups, or between
   individual samples?

:::



## Save data

```{r, include=FALSE}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson02.rda"))
```


::: {.callout-tip}
#### Key Points

- DIA-NN outputs a `.parquet` format report file containing precursor- and protein-level quantification data for each sample
- The `readQFeaturesFromDIANN` function imports DIA-NN output into a `QFeatures` object, creating one `SummarizedExperiment` per sample run
- Sample metadata is stored in the `colData` of the `QFeatures` object and can be used to annotate samples with experimental group information
- The `rowData` of each per-sample `SummarizedExperiment` contains quality metrics from DIA-NN (e.g. FWHM, Q-values) that can be used to explore data quality and guide downstream filtering decisions

:::

## References {-}
