---
title: Import and infrastructure
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Understand how DIA-NN output data is structured and how it can be imported into R
* Know how to use `readQFeaturesFromDIANN` to create a `QFeatures` object from DIA-NN output
* Be able to access sample metadata via `colData` in a `QFeatures` object
* Understand how to filter DIA-NN precursor data using Q-value thresholds

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("arrow")
library("readxl")
library("stringr")
library("QFeatures")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")
```


## Importing DIA-NN output data

The raw DIA mass spectrometry data was processed using DIA-NN, which outputs a
report file in `.parquet` format. This file contains precursor- and protein-level
identifications and quantities for each sample. We begin by reading this file
into R using the `read_parquet` function from the `arrow` package.

```{r}
diann_df <- read_parquet(here("course_files/dia/DIANNout/monkeypox_plasma_proteomes.parquet"))
```

### Cleaning run names

DIA-NN pastes together the name of the `.wiff2` folder and the `.wiff.scan`
files into the `Run` column. We need to clean up these run names so that they
match the `sample_id` field in our metadata.

```{r}
### Clean Run names
print(head(unique(diann_df$Run), 2))

# DIANN has pasted the name of the .wiff2 folder as well as the .wiff.scan files together.
# So we need to clean up the Run names so that they match the metadata `sample_id` field.
diann_df <- diann_df %>%
  mutate(Run = sub('\\..*', '', Run))

print(head(unique(diann_df$Run), 2))
```

## Importing sample metadata

We import the sample metadata from an Excel file. This provides information
about each sample, including which group (MPXV, COVID-19, or Control) each
run belongs to. We rename the `sample_id` column to `runCol` so it is
compatible with the `readQFeaturesFromDIANN` function.

```{r}
colData <- read_excel(here("course_files/dia/metadata/monkeypox_metadata.xlsx")) %>%
  dplyr::rename(runCol = sample_id)
```

## Reading into a `QFeatures` object

We use the `readQFeaturesFromDIANN` function to import the cleaned DIA-NN data
into a `QFeatures` object. This function creates one `SummarizedExperiment`
per run, with each assay containing the precursor-level quantification data
for that run. The `quantCols` argument specifies which column contains the
quantitative data, and `fnames` specifies the column to use as feature names.

```{r}
dia_qf <- QFeatures::readQFeaturesFromDIANN(
  diann_df, quantCols = "Precursor.Quantity", fnames = "Precursor.Id",
  runCol = 'Run', colData = colData)

colData(dia_qf)
```

## Exploring the data

### Full-width at half maximum (FWHM)

The `rowData` of each per-sample `SummarizedExperiment` contains quality
metrics from DIA-NN, including the Full-Width at Half Maximum (FWHM) of the
chromatographic peaks. We can extract this information across all runs and
visualise the distribution of FWHM values per sample, coloured by group.

```{r, fig.width=5, fig.height=5}
cdata <- data.frame(colData(dia_qf))

rdata <- rbindRowData(dia_qf, i = names(dia_qf)) %>%
  data.frame() %>%
  merge(cdata, by.x = "Run", by.y = "runCol")

rdata %>%
  ggplot(aes(FWHM, group = Run, colour = group)) +
  geom_density() +
  theme_classic()
```

### How many precursors per sample?

TS: POSSIBLE EXERCISE POINT - HOW MANY PRECURSORS FOR EACH SAMPLE

```{r, fig.height=4, fig.width=4}
# Look for suitable methods with ?QFeatures
hist(nrows(dia_qf))
```

## Filtering by Q-value

TS: INTERACTION POINT. WHY DO WE NEED TO FILTER? WHAT SHOULD WE FILTER ON? WHAT DO THESE COLUMNS MEAN?

DIA-NN outputs several Q-value columns that can be used to filter our data to
high-confidence identifications. The key Q-value columns are:

- **Q.Value** — run-level precursor confidence (false discovery rate)
- **PG.Q.Value** — run-level protein group confidence
- **Lib.Q.Value** — library-level precursor confidence (relevant when using match-between-runs)
- **Lib.PG.Q.Value** — library-level protein group confidence

Since our data was processed with match-between-runs (MBR) enabled, we
filter on the library-level Q-values (`Lib.Q.Value` and `Lib.PG.Q.Value`),
as these are most relevant for MBR-processed data. We apply a standard 1%
FDR threshold (Q-value ≤ 0.01) on all four columns.

```{r, fig.height=4, fig.width=4}

# From docs:
# Lib refers to the respective value saved in the spectral library, e.g. Lib.Q.Value means q-value for the respective library precursor; when using MBR, Lib q-values will correspond to q-values in the empirical DIA-based lib created in the first pass of MBR
# We should be filtering on Lib values here because these are the most relevant for our data, which was processed with MBR on.

dia_qf <- dia_qf %>%
  filterFeatures(~ Q.Value <= 0.01) %>%        # Run-level precursor Q.value
  filterFeatures(~ PG.Q.Value <= 0.01) %>%     # Run-level Protein group Q.value
  filterFeatures(~ Lib.Q.Value <= 0.01) %>%    # Library-level precursor Q.value
  filterFeatures(~ Lib.PG.Q.Value <= 0.01)     # Library-level Protein group Q-value

hist(nrows(dia_qf))
```

## Save data

```{r}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson02.rda"))
```


::: {.callout-tip}
#### Key Points

- DIA-NN outputs a `.parquet` format report file containing precursor- and protein-level quantification data for each sample
- The `readQFeaturesFromDIANN` function imports DIA-NN output into a `QFeatures` object, creating one `SummarizedExperiment` per sample run
- Sample metadata is stored in the `colData` of the `QFeatures` object and can be used to annotate samples with experimental group information
- DIA-NN Q-value columns (`Q.Value`, `PG.Q.Value`, `Lib.Q.Value`, `Lib.PG.Q.Value`) can be used to filter to high-confidence precursor and protein identifications at a desired FDR threshold

:::

## References {-}
