---
title: Data cleaning
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to use `filterFeatures` to filter DIA-NN precursor data by Q-value thresholds
* Know how to use `joinAssays` to combine per-sample `SummarizedExperiment` objects into a single combined set
* Be able to explore and visualise missing values in DIA proteomics data
* Understand how to filter sparse precursors based on a missing value threshold

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson02.rda"), verbose = TRUE)
```


Now that we have our precursor-level data stored in a `QFeatures` object, the
next step is to carry out data cleaning — filtering to retain only
high-confidence identifications and removing precursors with excessive missing
values across samples.

## Filtering by Q-value

::: {.callout-discussion}
#### Discussion

Before reading on, consider the following:

- Why do we need to filter the DIA-NN output at all? What would happen if we
  kept all identifications?
- What do the Q-value columns represent, and what is the difference between
  run-level and library-level Q-values?
- What threshold would you choose?
- What impact could this filtering have on the downstream analysis and results?

:::

DIA-NN assigns a Q-value to each precursor and protein group identification,
representing the local false discovery rate (FDR) at that score threshold —
i.e., the expected proportion of identifications at that level that are
incorrect. Filtering by Q-value ensures we retain only high-confidence
identifications before proceeding to downstream analysis.

DIA-NN outputs several Q-value columns:

- **Q.Value** — run-level precursor Q-value (FDR within a single run)
- **PG.Q.Value** — run-level protein group Q-value
- **Lib.Q.Value** — library-level precursor Q-value (based on the spectral
  library used in matching)
- **Lib.PG.Q.Value** — library-level protein group Q-value

Since our data was processed with match-between-runs (MBR) enabled, we
additionally filter on the library-level Q-values (`Lib.Q.Value` and
`Lib.PG.Q.Value`). When MBR is used, DIA-NN rescores identifications across
runs using an empirical library built in a first pass; the library-level
Q-values reflect the confidence of those cross-run matches and are therefore
the most stringent and appropriate filter for MBR data. We apply a standard
1% FDR threshold (Q-value ≤ 0.01) to all four columns.

{{< include ../_shared/_filterFeatures.md >}}

Since we want to apply the Q-value filter across all per-sample
`SummarizedExperiment`s simultaneously, we do not pass an `i` argument —
this causes `filterFeatures` to apply the filter to every set in the
`QFeatures` object.

```{r}
dia_qf

dia_qf <- dia_qf %>%
  filterFeatures(~ Q.Value <= 0.01) %>%       # Run-level precursor Q-value
  filterFeatures(~ PG.Q.Value <= 0.01) %>%    # Run-level protein group Q-value
  filterFeatures(~ Lib.Q.Value <= 0.01) %>%   # Library-level precursor Q-value
  filterFeatures(~ Lib.PG.Q.Value <= 0.01)    # Library-level protein group Q-value

dia_qf
```

We can see how many precursors remain per sample after filtering by inspecting
the number of rows in each per-sample set.

```{r, fig.height=4, fig.width=4}
hist(nrows(dia_qf),
     xlab = "Precursors",
     main = "Precursor counts per sample\nafter Q-value filtering")
```


## Joining per-sample sets into a single precursor set

After filtering, our `QFeatures` object contains one `SummarizedExperiment`
per sample run. To work with the data combined across all samples,
we use the `joinAssays` function. This merges all per-sample sets into a
single combined set called `"precursors"`, aligning features (precursors)
by their `Precursor.Id`. Where a precursor was not detected in a particular
sample, the corresponding entry will be `NA`.

```{r}
dia_qf <- joinAssays(x = dia_qf,
                     i = names(dia_qf),
                     name = "precursors",
                     fcol = "Precursor.Id")
```

Note that `joinAssays` only retains `rowData` columns that are shared across
all sets and have consistent values — run-specific metrics such as `Q.Value`,
`PEP`, and `Run` are therefore dropped from the combined set. This is
expected: these per-run quality metrics were used for filtering in the previous
step and are no longer needed. We can confirm which columns are dropped by
comparing the `rowData` names before and after joining.

```{r}
# rowData columns in the per-sample sets
sample_level_rdata_names <- rowDataNames(dia_qf)[[1]]

# rowData columns in the combined set
joined_rdata_names <- rowDataNames(dia_qf)[["precursors"]]

# Columns present in sample-level sets but dropped from the combined set
setdiff(sample_level_rdata_names, joined_rdata_names)
```

## Removing individual sample-level assays

Now that the per-sample data has been joined into the `"precursors"` set, we
no longer need to keep the individual per-sample `SummarizedExperiment` objects.
To reduce the size of our QFeatures object, we identify the sets for individual
samples and remove them. This isn't strictly necessary for this dataset, but it
may be if you have a very large experiment. It also helps to keep the object
tidy and focused on the combined set for downstream analysis.

```{r}
sets_to_rm <- which(names(dia_qf) != "precursors")
dia_qf <- removeAssay(dia_qf, i = sets_to_rm)
```


## Exploring missing values

DIA proteomics data can exhibit substantial missing values because proteins must
be detected independently in each sample run. Before filtering, it is important
to understand the extent and distribution of missing values in the data.

We use the `nNA` function to calculate the proportion of missing values
per precursor (row) and per sample (column).

```{r}
precursors_missing <- nNA(dia_qf, i = "precursors")

print(precursors_missing)
```

:::{.callout-exercise}
#### Challenge: Visualising missing values

{{< level 2 >}}

Using the `precursors_missing` object, create plots to explore the
distribution of missing values in the data.

1. Plot a histogram for the proportion of missing values across _precursors_.
2. Plot the proportion of missing values per _sample_, coloured by _group_.

Based on your plots:

- Do most precursors have missing values?
- Are there any precursors or samples with an unusually high proportion of
  missing values?
- Does the proportion of missing values differ between groups?


<details><summary>Hint</summary>The `precursors_missing` object contains
two data frames: `nNArows` (one row per precursor) and `nNAcols` (one row
per sample), each with a `pNA` column giving the proportion of missing
values.</details>

::: {.callout-answer collapse="true"}

**Task 1**

```{r}
hist(precursors_missing$nNArows$pNA,
     xlab = "Proportion missing values",
     main = "Missing values per precursor")
```

**Task 2**

```{r}
precursors_missing$nNAcols %>%
  as_tibble() %>%
  merge(data.frame(colData(dia_qf)), by.x = "name", by.y = "runCol") %>%
  ggplot(aes(x = name, y = pNA, group = group, fill = group)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Proportion missing values") +
  coord_flip() +
  scale_fill_brewer(palette = "Dark2") +
  theme_classic()
```

:::

:::

Precursors that are missing across the majority of samples provide little
useful quantitative information and may introduce noise into downstream
analyses. We filter out precursors with excessive missing values using the
`filterNA` function, where the `pNA` argument specifies the maximum proportion
of missing values allowed per precursor.

The choice of threshold involves a trade-off: stricter thresholds (lower `pNA`)
remove more precursors and produce a cleaner but smaller dataset; more
permissive thresholds retain more precursors but include those with very sparse
quantification, which are more likely to be inaccurately quantified. Here we use
`pNA = 0.75`, meaning a precursor must be observed in at least 25% of samples —
with 31 samples, this corresponds to a minimum of approximately 8 observations.
Further filtering will be applied at the protein level in the statistical
analysis, where we will require proteins to be quantified in at least 3
replicates per group before testing.

## Filtering sparse precursors

{{< include ../_shared/_copy_assay_rationale.md >}}

We call the copy `"precursors_filtered_missing"` — after filtering, this set
will contain only precursors that meet the missing value threshold. We use
`plot` to visualise the assay structure before and after adding the link.

```{r, fig.height=3, fig.width=3}
dia_qf <- addAssay(dia_qf, dia_qf[['precursors']],
                   name = "precursors_filtered_missing")

plot(dia_qf) # no link yet

dia_qf <- addAssayLink(dia_qf, from = "precursors",
                       to = "precursors_filtered_missing")

plot(dia_qf) # link now shown
```

Now we perform the missing value filtering on the new assay, specifying the `i` argument to apply the filter only to the `"precursors_filtered_missing"` assay. This retains the original `"precursors "` assay with all precursors for reference, while the new set contains only those precursors that pass the missing value threshold.

```{r}
dia_qf <- dia_qf %>%
  filterNA(pNA = 0.75, i = "precursors_filtered_missing")
```

TS: Update data to contaminants-aware search and include a section on filtering out contaminants and decoys, as well as a discussion of how to handle them in downstream analysis, as per the TMT notebook 3.

## Save data

```{r, include=FALSE}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson03.rda"))
```


::: {.callout-tip}
#### Key Points

- DIA-NN Q-value columns (`Q.Value`, `PG.Q.Value`, `Lib.Q.Value`,
  `Lib.PG.Q.Value`) represent the local FDR for each identification; filtering
  at ≤ 0.01 retains only identifications at 1% FDR
- When match-between-runs is enabled, library-level Q-values (`Lib.Q.Value`,
  `Lib.PG.Q.Value`) are the most appropriate filter as they reflect the
  confidence of cross-run matches
- `filterFeatures` applies row-level filters based on `rowData` variables,
  and can be applied to the whole `QFeatures` object or to specific sets
- `joinAssays` combines per-sample `SummarizedExperiment` objects into a single
  combined set, with missing values (`NA`) where a precursor was not detected
  in a given sample
- `nNA` provides a convenient summary of missing values per feature (row) and
  per sample (column)
- DIA data typically has a higher proportion of missing values than TMT-labelled
  data, as proteins must be detected independently in each sample
- `filterNA` removes precursors that are missing above a specified proportion
  of samples, with the threshold chosen based on the experimental context

:::

## References {-}
