---
title: Data processing
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to use `filterFeatures` to filter DIA-NN precursor data by Q-value thresholds
* Know how to use `joinAssays` to combine per-sample `SummarizedExperiment` objects into a single combined assay
* Be able to explore and visualise missing values in DIA proteomics data
* Understand how to filter sparse precursors based on a missing value threshold

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson02.rda"), verbose = TRUE)
```


## Filtering by Q-value

::: {.callout-discussion}
#### Discussion

Before reading on, consider the following:

- Why do we need to filter the DIA-NN output at all? What would happen if we
  kept all identifications?
- What do the Q-value columns represent, and what is the difference between
  run-level and library-level Q-values?
- What threshold would you choose?
- What impact could this filtering have on the downstream analysis and results?

:::

DIA-NN assigns a Q-value to each precursor and protein group identification,
representing the local false discovery rate (FDR) at that score threshold —
i.e., the expected proportion of identifications at that level that are
incorrect. Filtering by Q-value ensures we retain only high-confidence
identifications before proceeding to downstream analysis.

DIA-NN outputs several Q-value columns:

- **Q.Value** — run-level precursor Q-value (FDR within a single run)
- **PG.Q.Value** — run-level protein group Q-value
- **Lib.Q.Value** — library-level precursor Q-value (based on the spectral
  library used in matching)
- **Lib.PG.Q.Value** — library-level protein group Q-value

Since our data was processed with match-between-runs (MBR) enabled, we
additionally filter on the library-level Q-values (`Lib.Q.Value` and
`Lib.PG.Q.Value`). When MBR is used, DIA-NN rescores identifications across
runs using an empirical library built in a first pass; the library-level
Q-values reflect the confidence of those cross-run matches and are therefore
the most stringent and appropriate filter for MBR data. We apply a standard
1% FDR threshold (Q-value ≤ 0.01) to all four columns.

{{< include ../_shared/_filterFeatures.md >}}

Since we want to apply the Q-value filter across all per-sample
`SummarizedExperiment`s simultaneously, we do not pass an `i` argument —
this causes `filterFeatures` to apply the filter to every set in the
`QFeatures` object.

```{r}
dia_qf

dia_qf <- dia_qf %>%
  filterFeatures(~ Q.Value <= 0.01) %>%       # Run-level precursor Q-value
  filterFeatures(~ PG.Q.Value <= 0.01) %>%    # Run-level protein group Q-value
  filterFeatures(~ Lib.Q.Value <= 0.01) %>%   # Library-level precursor Q-value
  filterFeatures(~ Lib.PG.Q.Value <= 0.01)    # Library-level protein group Q-value

dia_qf
```

We can see how many precursors remain per sample after filtering by inspecting
the number of rows in each per-sample assay.

```{r, fig.height=4, fig.width=4}
hist(nrows(dia_qf),
     xlab = "Precursors",
     main = "Precursor counts per sample after Q-value filtering")
```


## Joining per-sample assays into a single precursor matrix

After filtering, our `QFeatures` object contains one `SummarizedExperiment`
per sample run. To work with the data as a combined matrix across all samples,
we use the `joinAssays` function. This merges all per-sample assays into a
single combined assay called `"precursors"`, aligning features (precursors)
by their `Precursor.Id`. Where a precursor was not detected in a particular
sample, the corresponding entry in the combined matrix will be `NA`.

```{r}
dia_qf <- joinAssays(x = dia_qf,
                     i = names(dia_qf),
                     name = "precursors",
                     fcol = "Precursor.Id")
```

Note that `joinAssays` only retains `rowData` columns that are shared across
all assays and have consistent values — run-specific metrics such as `Q.Value`,
`PEP`, and `Run` are therefore dropped from the combined assay. This is
expected: these per-run quality metrics were used for filtering in the previous
step and are no longer needed. We can confirm which columns are dropped by
comparing the `rowData` names before and after joining.

```{r}
# rowData columns in the per-sample assays
sample_level_rdata_names <- rowDataNames(dia_qf)[[1]]

# rowData columns in the combined assay
joined_rdata_names <- rowDataNames(dia_qf)[["precursors"]]

# Columns present in sample-level assays but dropped from the combined assay
setdiff(sample_level_rdata_names, joined_rdata_names)
```

## Removing individual sample-level assays

Now that the per-sample data has been joined into the `"precursors"` assay,
we no longer need to keep the individual per-sample `SummarizedExperiment`
objects. We identify and remove them to keep our `QFeatures` object tidy.

```{r}
sets_to_rm <- which(names(dia_qf) != "precursors")
dia_qf <- removeAssay(dia_qf, i = sets_to_rm)
```




## Exploring missing values

DIA proteomics data can exhibit substantial missing values because proteins must
be detected independently in each sample run. Before filtering, it is important
to understand the extent and distribution of missing values in the data.

We use the `nNA` function to calculate the proportion of missing values
per precursor (row) and per sample (column).

```{r}
precursors_missing <- nNA(dia_qf, i = "precursors")

print(precursors_missing)
```

:::{.callout-exercise}
#### Challenge: Visualising missing values

{{< level 2 >}}

Using the `precursors_missing` object, create plots to explore the
distribution of missing values in the data.

1. Plot the distribution of the proportion of missing values across precursors.
2. Plot the proportion of missing values per sample, coloured by group.

Based on your plots:

- Does the proportion of missing values differ between groups?
- Are there any precursors or samples with an unusually high proportion of
  missing values?

<details><summary>Hint</summary>The `precursors_missing` object contains
two data frames: `nNArows` (one row per precursor) and `nNAcols` (one row
per sample), each with a `pNA` column giving the proportion of missing
values.</details>

::: {.callout-answer collapse="true"}

**Task 1**

```{r}
hist(precursors_missing$nNArows$pNA,
     xlab = "Proportion missing values",
     main = "Missing values per precursor")
```

**Task 2**

```{r}
precursors_missing$nNAcols %>%
  as_tibble() %>%
  merge(data.frame(colData(dia_qf)), by.x = "name", by.y = "runCol") %>%
  ggplot(aes(x = name, y = pNA, group = group, fill = group)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Proportion missing values") +
  coord_flip() +
  scale_fill_brewer(palette = "Dark2") +
  theme_classic()
```

:::

:::

## Filtering sparse precursors

Precursors that are missing across the majority of samples provide little
useful quantitative information and may introduce noise into downstream
analyses. We filter out precursors with excessive missing values using the
`filterNA` function with a threshold of `pNA = 0.75`, meaning precursors
missing in more than 75% of samples are removed.

We first create a copy of the precursors assay to retain the unfiltered data,
and add an assay link to maintain the relationship between the two assays.

```{r, fig.height=3, fig.width=3}
dia_qf <- addAssay(dia_qf, dia_qf[['precursors']],
                   name = "precursors_filtered_missing")

plot(dia_qf) # no automatic link between assays

dia_qf <- addAssayLink(dia_qf, from = "precursors",
                       to = "precursors_filtered_missing")

plot(dia_qf)
```

```{r}
dia_qf <- dia_qf %>%
  filterNA(pNA = 0.75, i = "precursors_filtered_missing")
```

## Save data

```{r}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson03.rda"))
```


::: {.callout-tip}
#### Key Points

- DIA-NN Q-value columns (`Q.Value`, `PG.Q.Value`, `Lib.Q.Value`,
  `Lib.PG.Q.Value`) represent the local FDR for each identification; filtering
  at ≤ 0.01 retains only identifications at 1% FDR
- When match-between-runs is enabled, library-level Q-values (`Lib.Q.Value`,
  `Lib.PG.Q.Value`) are the most appropriate filter as they reflect the
  confidence of cross-run matches
- `filterFeatures` applies row-level filters based on `rowData` variables,
  and can be applied to the whole `QFeatures` object or to specific sets
- `joinAssays` combines per-sample `SummarizedExperiment` objects into a single
  combined assay, with missing values (`NA`) where a precursor was not detected
  in a given sample
- `nNA` provides a convenient summary of missing values per feature (row) and
  per sample (column)
- DIA data typically has a higher proportion of missing values than TMT-labelled
  data, as proteins must be detected independently in each sample
- `filterNA` removes precursors that are missing above a specified proportion
  of samples, with the threshold chosen based on the experimental context

:::

## References {-}
