---
title: Data processing
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to use `joinAssays` to combine per-sample `SummarizedExperiment` objects into a single combined assay
* Be able to explore and visualise missing values in DIA proteomics data
* Understand how to filter sparse precursors based on a missing value threshold

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("arrow")
library("readxl")
library("stringr")
library("QFeatures")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson02.rda"), verbose = TRUE)
```


## Joining per-sample assays into a single precursor matrix

After filtering, our `QFeatures` object contains one `SummarizedExperiment`
per sample run. To work with the data as a combined matrix across all samples,
we use the `joinAssays` function. This merges all per-sample assays into a
single combined assay called `"precursors"`, aligning features (precursors)
by their `Precursor.Id`.

```{r}
# make combined precursor SE
dia_qf <- joinAssays(x = dia_qf,
                     i = names(dia_qf),
                     name = "precursors",
                     fcol = "Precursor.Id")
```

## Removing individual sample-level assays

Now that the per-sample data has been joined into the `"precursors"` assay,
we no longer need to keep the individual per-sample `SummarizedExperiment`
objects. We identify and remove them to keep our `QFeatures` object tidy.

```{r}
## Identify which sets to remove
sets_to_rm <- which(names(dia_qf) != "precursors")
## Now create a new object and remove them
dia_qf <- removeAssay(dia_qf, i = sets_to_rm)
```

## Exploring missing values

DIA proteomics data can exhibit missing values because proteins must be
detected independently in each sample. Before filtering, it is important to
understand the extent and distribution of missing values in the data.

We use the `nNA` function to calculate the proportion of missing values
per precursor (row) and per sample (column).

```{r}
precursors_missing <- nNA(dia_qf, i = "precursors")

print(precursors_missing)
```

TS: POSSIBLE EXCERCISE POINT - PLOT MISSING VALUES. ASK Q: IS THIS OK? DO WE NEED TO FILTER OUT SOME PRECURSORS? DO WE NEED TO FILTER OUT ANY SAMPLES?

The histogram below shows the distribution of the proportion of missing values
per precursor. The barplot shows the proportion of missing values per sample,
coloured by experimental group.

```{r}
hist(precursors_missing$nNArows$pNA)

precursors_missing$nNAcols %>%
  as_tibble() %>%
  merge(data.frame(colData(dia_qf)), by.x = "name", by.y = "runCol") %>%
  ggplot(aes(x = name, y = pNA, group = group, fill = group)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Proportion missing values") +
  coord_flip() +
  scale_fill_brewer(palette = "Dark2") +
  theme_classic()
```

## Filtering sparse precursors

Precursors that are missing across the majority of samples provide little
useful quantitative information and may introduce noise into downstream
analyses. We filter out precursors with excessive missing values using the
`filterNA` function with a threshold of `pNA = 0.75`, meaning precursors
missing in more than 75% of samples are removed.

We first create a copy of the precursors assay to retain the unfiltered data,
and add an assay link to maintain the relationship between the two assays.

```{r, fig.height=3, fig.width=3}
dia_qf <- addAssay(dia_qf, dia_qf[['precursors']],
                   name = "precursors_filtered_missing")

plot(dia_qf) # no automatic link between assays

dia_qf <- addAssayLink(dia_qf, from = "precursors",
                       to = "precursors_filtered_missing")

plot(dia_qf)
```

Filtering by missing values using filterNA

```{r}
dia_qf <- dia_qf %>%
  filterNA(pNA = 0.75, i = "precursors_filtered_missing")
```

## Save data

```{r}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson03.rda"))
```


::: {.callout-tip}
#### Key Points

- `joinAssays` combines per-sample `SummarizedExperiment` objects into a single combined assay, enabling analysis across all samples simultaneously
- `nNA` provides a convenient summary of missing values per feature (row) and per sample (column)
- DIA data typically has a higher proportion of missing values than TMT-labelled data, as proteins must be detected independently in each sample
- `filterNA` can be used to remove precursors that are missing above a specified proportion of samples, with the threshold chosen based on the experimental context

:::

## References {-}
