---
title: Normalisation and aggregation
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to log2 transform DIA precursor data within a `QFeatures` object
* Understand how to use `robustSummary` to summarise precursor-level data to protein-level abundance, including in the presence of missing values
* Be able to normalise protein-level data using the `diff.median` method
* Understand the importance of visualising missing values at the protein level

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("naniar")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson03.rda"), verbose = TRUE)
```


Having filtered our DIA-NN data to retain only high-confidence identifications
and removed precursors with excessive missing values, we are now ready to
continue processing. In this lesson we will log2 transform the precursor-level
data, aggregate to protein-level abundances, and normalise the protein-level
data ready for downstream statistical analysis.

## Logarithmic transformation

If we look at the distribution of raw precursor intensities, we will see that
the abundance values are dramatically skewed towards zero.

```{r, warning=FALSE, message=FALSE}
dia_qf[["precursors_filtered_missing"]] %>%
  assay() %>%
  longForm() %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  theme_bw() +
  xlab("Abundance (raw)")
```

{{< include ../_shared/_log_transformation.md >}}

::: {.callout-note}
#### At which stage should log transformation be applied?
Logarithmic transformation should be applied at the appropriate stage of the
processing pipeline for the aggregation method being used. Here we apply it
**before** aggregation to protein level, because we use `robustSummary` for
summarisation. `robustSummary` fits a linear model to the precursor-level data,
which assumes approximately Gaussian distributed residuals — a requirement that
is only met after log transformation.
:::

We apply a log2 transformation to the filtered precursor data using the
`logTransform` function, which creates a new set in our `QFeatures` object.

```{r}
dia_qf <- logTransform(
  dia_qf,
  base = 2,
  i = "precursors_filtered_missing",
  name = "precursors_filtered_missing_log")
```

## Summarising to protein-level abundance

We now aggregate from precursor level to protein level using the
`aggregateFeatures` function. For DIA data with missing values, we use
the `robustSummary` method (@Sticker2020), which models the log-transformed
peptide/precursor-level quantification as being dependent upon the
protein-level abundance plus a precursor-level effect. This modelling-based
approach can handle missing values without imputation, as it only considers
finite data. The only requirement is that a precursor must be quantified in
at least two samples.

```{r qfeatures-aggregate-precursors-to-proteins, message=FALSE}
dia_qf <- aggregateFeatures(dia_qf,
                            i = "precursors_filtered_missing_log",
                            fcol = "Protein.Ids",
                            name = "proteins",
                            fun = MsCoreUtils::robustSummary,
                            maxit = 10000)

dia_qf
```

:::{.callout-exercise}
#### Challenge: Quantification completeness at the protein level

{{< level 2 >}}

Although `robustSummary` can handle missing values at the precursor level, a
protein will still have a missing abundance in any sample where all of its
precursors are absent.

To investigate this, we will use the `longForm` function, which converts a
`QFeatures` set from wide format (proteins × samples) to long format (one
row per protein–sample combination). The `colvars` argument specifies which
columns from the `colData` to carry across.

Run the following code and inspect the output — what does each row represent?

```{r}
# dia_qf[,,'proteins'] subsets the QFeatures object to the "proteins" set
longForm(dia_qf[,,'proteins'], colvars = 'group') %>%
  data.frame() %>%
  head()
```

Now use this long-format data to create a plot showing how many samples each
protein is quantified in, broken down by group. Then consider:

- Why might some proteins still be missing in some samples despite using
  `robustSummary`?
- What could you do about proteins with extensive missingness — impute,
  filter, or ignore?

<details><summary>Hint</summary>Count the number of finite values per protein
per group using `sum(is.finite(value))`.</details>

::: {.callout-answer collapse="true"}

Each row in the long-format data represents one protein–sample observation,
with `value` containing the protein abundance and `group` indicating which
experimental group that sample belongs to.

```{r}
longForm(dia_qf[,,'proteins'], colvars = 'group') %>%
  data.frame() %>%
  group_by(rowname, group) %>%
  summarise(n_quant = sum(is.finite(value))) %>%
  ggplot(aes(n_quant)) +
  geom_histogram() +
  facet_wrap(~group) +
  theme_classic() +
  labs(x = 'Quantified samples', y = 'Proteins')
```

`robustSummary` estimates protein abundance from whatever precursors are
observed in a given sample. If *no* precursors for a protein are observed in
that sample, there is no data from which to estimate abundance, so the
protein-level value remains missing regardless of the summarisation method.

:::

:::


:::{.callout-exercise}
#### Challenge: Patterns of missingness

{{< level 3 >}}

The previous exercise showed *how many* samples each protein is quantified
in, but not *which* samples. Understanding the pattern of missingness —
whether it is random or associated with particular samples or groups — can
inform decisions about how to handle missing values downstream.

Use the `naniar::gg_miss_upset` function to visualise the patterns of
missingness in the protein-level data. An upset plot shows which combinations
of samples tend to have missing values for the same proteins.

- Are missing values random across samples, or do they tend to cluster within
  particular groups?
- What does this tell you about the likely cause of the missingness — is it
  more consistent with MCAR or MNAR?

<details><summary>Hint</summary>Extract the protein assay matrix with
`assay()`, convert to a data frame, and pass to `naniar::gg_miss_upset`.
Each column (sample) becomes a "set" in the upset plot. Use
`sets = paste0(colnames(missing_data), "_NA")` to control the order of
samples, and `nintersects` to limit the number of intersections
shown.</details>

::: {.callout-answer collapse="true"}

```{r, fig.height=10, fig.width=10, warning=FALSE}
missing_data <- dia_qf[['proteins']] %>% assay() %>% data.frame()

# The sets argument is required to make the keep_order work as intended,
# otherwise the order of the sets is determined by the total number of missing
# values in each set, which is not what we want here. By providing the sets
# argument with the column names of the missing data, we can ensure that the order
# of the sets in the upset plot matches the order of the columns in our data frame.
naniar::gg_miss_upset(missing_data,
                      sets = paste0(colnames(missing_data), "_NA"),
                      keep.order = TRUE,
                      nintersects = 50)
```

If missing values are clustered within specific groups rather than randomly
distributed across samples, this suggests the missingness is
intensity-dependent (MNAR) — proteins that are genuinely absent or below the
limit of detection in certain conditions. This is biologically informative
but has implications for imputation: left-censored methods are more
appropriate for MNAR data than methods that assume values are missing at
random.

:::

:::

## Normalisation

{{< include ../_shared/_normalisation_intro.md >}}

{{< include ../_shared/_median_normalisation.md >}}

We apply the `diff.median` method, which shifts each sample's intensity
distribution so that all sample medians match the grand median across all
samples.

::: {.callout-note}
#### Alternative normalisation methods
More sophisticated normalisation approaches may be appropriate depending on the
data. For example, the authors of the original study from which this dataset is
derived [@Wang2022] used cyclic loess normalisation, available via
`limma::normalizeCyclicLoess()`, which corrects for intensity-dependent biases
rather than applying a simple global shift. However, applying it within the
QFeatures pipeline requires extracting and replacing the assay matrix manually,
and because cyclic loess is sensitive to missing values, additional care is
needed — complexity beyond the scope of this course.
:::

```{r qfeatures-normalize}
dia_qf <- normalize(dia_qf,
                    i = "proteins",
                    name = "norm_proteins",
                    method = "diff.median")
```

We can visualise the effect of normalisation by plotting density curves of
protein abundance per sample before and after normalisation.

```{r, warning=FALSE}
pre_norm <- longForm(dia_qf[,,'proteins'], colvars = 'group') %>%
  ggplot(aes(x = value, colour = group, group = colname)) +
  geom_density() +
  theme_classic() +
  xlab("log2 (Abundance)") +
  ggtitle('Pre-normalisation')

post_norm <- longForm(dia_qf[,,'norm_proteins'], colvars = 'group') %>%
  ggplot(aes(x = value, colour = group, group = colname)) +
  geom_density() +
  theme_classic() +
  xlab("log2 (Abundance)") +
  ggtitle('Post-normalisation')

pre_norm + post_norm
```

## Save data

```{r, include=FALSE}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson04.rda"))
```


::: {.callout-tip}
#### Key Points

- Log2 transformation is applied to precursor-level data before aggregation to protein level, giving data a more Gaussian distribution suitable for downstream statistical analysis
- `robustSummary` is the recommended summarisation method for DIA data, as it can handle missing values at the precursor level without requiring imputation
- Proteins where all precursors are missing in a given sample will still have missing protein-level abundances, even with `robustSummary` — these should be explored and handled appropriately
- Upset plots are a useful way to visualise patterns of missingness across samples, helping to identify whether missing values are random or associated with particular groups
- `diff.median` normalisation corrects for sample-level systematic differences by shifting each sample's intensity distribution to have the same median

:::

## References {-}
