---
title: Normalisation and aggregation
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to log2 transform DIA precursor data within a `QFeatures` object
* Understand how to use `robustSummary` to summarise precursor-level data to protein-level abundance, including in the presence of missing values
* Be able to normalise protein-level data using the `diff.median` method
* Understand the importance of visualising missing values at the protein level

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("arrow")
library("readxl")
library("stringr")
library("QFeatures")
library("naniar")
library("tidyr")
library("dplyr")
library("ggplot2")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson03.rda"), verbose = TRUE)
```


## Logarithmic transformation

{{< include ../_shared/_log_transformation.md >}}

We apply a log2 transformation to the filtered precursor data using the
`logTransform` function, which creates a new assay in our `QFeatures` object.

```{r}
dia_qf <- logTransform(
  dia_qf,
  base = 2,
  i = "precursors_filtered_missing",
  name = "precursors_filtered_missing_log")
```

## Summarising to protein-level abundance

We now aggregate from precursor level to protein level using the
`aggregateFeatures` function. For DIA data with missing values, we use
the `robustSummary` method (@Sticker2020), which models the log-transformed
peptide/precursor-level quantification as being dependent upon the
protein-level abundance plus a precursor-level effect. This modelling-based
approach can handle missing values without imputation, as it only considers
finite data. The only requirement is that a precursor must be quantified in
at least two samples.

```{r qfeatures-aggregate-precursors-to-proteins, message=FALSE}
dia_qf <- aggregateFeatures(dia_qf,
                            i = "precursors_filtered_missing_log",
                            fcol = "Protein.Ids",
                            name = "proteins",
                            fun = MsCoreUtils::robustSummary,
                            maxit = 10000)

dia_qf
```

TS: POSSIBLE EXCERCISE POINT - HOW MANY SAMPLES ARE QUANTIFIED FOR EACH PROTIEN. ASK Q: WHY ARE SOME SAMPLES NOT QUANTIFIED IF ROBUSTSUMMARY CAN HANDLE MISSING DATA? WHAT SHOULD WE DO ABOUT THIS - IMPUTE/FILTER/IGNORE?

Although `robustSummary` can handle missing values at the precursor level,
proteins where all precursors are missing in a given sample will still result
in a missing protein-level abundance for that sample. We can visualise how
many samples are quantified for each protein, broken down by experimental group.

```{r}
longForm(dia_qf[,,'proteins'], colvars = 'group') %>%
  data.frame() %>%
  group_by(rowname, group) %>%
  summarise(n_quant = sum(is.finite(value))) %>%
  ggplot(aes(n_quant)) +
  geom_histogram() +
  facet_wrap(~group) +
  theme_classic() +
  labs(x = 'Quantified samples', y = 'Proteins')
```


TS: POSSIBLE ((DIFFICULT) EXCERCISE POINT - WHAT ARE THE PATTERNS OF MISSINGNESS? WHAT DOES THIS TELL YOU ABOUT WHY SOME PROTEINS ARE MISSING. HOW DOES THIS AFFECT WHAT YOU DO NEXT?

TS: I'D START WITH AN UPSET PLOT TO SHOW THE PATTERNS OF MISSINGNESS ACROSS SAMPLES. THIS CAN BE DONE USING `naniar::::gg_miss_upset`. ALTERNATIVE WOULD BE TO PLOT A HEATMAP OF MISSINGNESS. THE UPSET PLOT IS PROBABLY THE MOST INFORMATIVE AND EASIEST TO INTERPRET, BUT IT DOES REQUIRE SOME EXPLANATION OF HOW TO READ IT.

```{r, fig.height=10, fig.width=10, warning=FALSE}
missing_data <- dia_qf[['proteins']] %>% assay() %>% data.frame()

# The sets argument is required to make the keep_order work as intended,
# otherwise the order of the sets is determined by the total number of missing
# values in each set, which is not what we want here. By providing the sets
# argument with the column names of the missing data, we can ensure that the order
# of the sets in the upset plot matches the order of the columns in our data frame.
naniar::gg_miss_upset(missing_data,
                      sets = paste0(colnames(missing_data), "_NA"),
                      keep.order = TRUE,
                      nintersects = 50)
```

## Normalisation

{{< include ../_shared/_normalisation_intro.md >}}

We normalise the protein-level data using the `diff.median` method, which
shifts the intensity distribution of each sample so that the median intensity
is the same across all samples. This corrects for sample-level systematic
differences in loading or ionisation efficiency.

```{r qfeatures-normalize}
dia_qf <- normalize(dia_qf,
                    i = "proteins",
                    name = "norm_proteins",
                    method = "diff.median")
```

We can visualise the effect of normalisation by plotting density curves of
protein abundance per sample before and after normalisation.

```{r}
pre_norm <- longForm(dia_qf[,,'proteins'], colvars = 'group') %>%
  ggplot(aes(x = value, colour = group, group = colname)) +
  geom_density() +
  theme_classic() +
  xlab("log2 (Abundance)") +
  ggtitle('Pre-normalisation')

post_norm <- longForm(dia_qf[,,'norm_proteins'], colvars = 'group') %>%
  ggplot(aes(x = value, colour = group, group = colname)) +
  geom_density() +
  theme_classic() +
  xlab("log2 (Abundance)") +
  ggtitle('Post-normalisation')

pre_norm + post_norm
```

## Save data

```{r}
save(dia_qf, file = here("course_files/dia/preprocessed/lesson04.rda"))
```


::: {.callout-tip}
#### Key Points

- Log2 transformation is applied to precursor-level data before aggregation to protein level, giving data a more Gaussian distribution suitable for downstream statistical analysis
- `robustSummary` is the recommended summarisation method for DIA data, as it can handle missing values at the precursor level without requiring imputation
- Proteins where all precursors are missing in a given sample will still have missing protein-level abundances, even with `robustSummary` — these should be explored and handled appropriately
- Upset plots are a useful way to visualise patterns of missingness across samples, helping to identify whether missing values are random or associated with particular groups
- `diff.median` normalisation corrects for sample-level systematic differences by shifting each sample's intensity distribution to have the same median

:::

## References {-}
