---
title: Exploration and visualisation of protein data
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Be able to visualise protein abundance for specific proteins of interest across experimental groups
* Appreciate the use of Principal Component Analysis (PCA) for visualising key factors that contribute to sample variation
* Complete PCA using the `prcomp` function from the `stats` package
* Be able to use the `subsetByFeature` function to get data across all levels for a feature of interest

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("limma")
library("factoextra")
library("tidyr")
library("dplyr")
library("tibble")
library("ggplot2")
library("ggrepel")
library("ggbeeswarm")
library("patchwork")

load(here("course_files/dia/preprocessed/lesson04.rda"), verbose = TRUE)
```

TS: MOVE THE PROTEINS OF INTEREST PLOT AND PCA TO DATA EXPLORATION SECTION

TS: ADD SECTION USING subsetByFeature FUNCTION HERE, AS PER THE TMT MATERIAL

## Plotting quantification values for proteins of interest

Often we are interested in the quantities of specific proteins, for example,
proteins we expect to change in abundance based on prior biological knowledge.
We can plot the quantification values for these proteins to inspect them
directly before any formal statistical testing.

We first convert the protein-level data to long format using `longForm`,
extracting both sample-level metadata (group, shortname) and feature-level
metadata (gene names, protein names).

```{r}
long_form_protein <- longForm(dia_qf[,,"norm_proteins"],
                              colvars = c('shortname', 'group'),
                              rowvars = c('Genes', 'Protein.Names')) %>%
  data.frame() %>%
  dplyr::rename(UniprotID = rowname)
```

We can then filter to a set of proteins of interest and plot abundance
values per sample, grouped by condition.

```{r, fig.height=6, fig.width=8}
proteins_of_interest <- c('P02766', # TTR
                          'P18428',  # LBP
                          'P13796') # LCP1

long_form_protein %>%
  # filter to the proteins of interest
  filter(UniprotID %in% proteins_of_interest) %>%
  # define aesthetics
  ggplot(aes(x = group, y = value, fill = group)) +
  # Using quasirandom from ggbeeswarm to avoid overlapping points
  geom_quasirandom(dodge.width = 0.5, colour = 'grey', pch = 21, size = 2) +
  # Summarise with mean +/- 95% confidence interval
  stat_summary(fun.data = mean_cl_normal, position = position_dodge(width = 0.5), show.legend = FALSE) +
  theme_classic() +
  labs(x = '', y = 'Protein abundance (log2)') +
  facet_wrap(~Genes, scales = 'free')
```

## Principal Component Analysis (PCA)

{{< include ../_shared/_pca.md >}}

### Completing PCA with `prcomp`

TS: INTERACTION POINT: DISCUSS MISSING VALUES HERE

Since PCA requires a complete data matrix, we first remove proteins with
any missing values using `filterNA`. We then extract the normalised
protein-level assay, transpose so that samples are rows, and pass to `prcomp`.

```{r}
protein_pca <- dia_qf[["norm_proteins"]] %>%
  filterNA() %>%
  assay() %>%
  t() %>%
  prcomp(scale = TRUE, center = TRUE)

summary(protein_pca)
```

We plot a scree plot to inspect how much variance each PC explains.

```{r}
fviz_screeplot(protein_pca)
```

TS: INTERACTION POINT: HOW WOULD YOU INTERPRET THESE PCAS?

We then plot the samples in PCA space, colouring by group and using shape
to indicate age group.

```{r}
p <- protein_pca$x %>%
  merge(colData(dia_qf), by = 'row.names') %>%
  data.frame() %>%
  ggplot(aes(x = PC1, y = PC2, colour = group, shape = age.group)) +
  geom_point(size = 3) +
  theme_bw()

print(p)
print(p + aes(x = PC3, y = PC4))
```

## The `subsetByFeature` function

TS: ADD SECTION USING subsetByFeature FUNCTION HERE, AS PER THE TMT MATERIAL

As well as visualising all proteins, it can be useful to extract all data
levels corresponding to a specific protein of interest. The `QFeatures`
infrastructure provides the `subsetByFeature` function to do this. It takes
a `QFeatures` object and one or more feature names, and returns a new
`QFeatures` object containing only data corresponding to those features.

Let's look at TTR (Transthyretin; UniProt ID P02766), one of the proteins
of interest plotted above.

```{r}
TTR <- subsetByFeature(dia_qf, "P02766")

experiments(TTR)
```

We can use this subset to visualise how the precursor data was aggregated
to protein level for this particular feature.

```{r, warning=FALSE, message=FALSE}
TTR[, , c("precursors_filtered_missing", "proteins")] %>%
  longForm() %>%
  as_tibble() %>%
  mutate(assay_order = factor(assay,
                              levels = c("precursors_filtered_missing",
                                         "proteins"))) %>%
  ggplot(aes(x = colname, y = value, colour = assay)) +
  geom_point() +
  geom_line(aes(group = rowname)) +
  theme(axis.text.x = element_text(angle = 45, size = 7)) +
  facet_wrap(~ assay_order)
```


::: {.callout-tip}
#### Key Points

- Plotting protein abundances for specific proteins of interest is a useful way to inspect the data before formal statistical testing
- PCA must be run on a complete data matrix; proteins with missing values in any sample should be removed with `filterNA` before applying `prcomp`
- In DIA data, `filterNA` prior to PCA may remove a substantial proportion of proteins due to the higher rate of missing values compared to TMT
- Colouring PCA plots by all potential sources of variation (e.g., group, age group) helps identify which factors drive the observed structure in the data
- The `subsetByFeature` function provides a convenient way to extract all data levels for a protein of interest from a `QFeatures` object

:::

## References {-}
