---
title: Visualisation of differential abundance results
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to produce heatmaps of protein abundance data using `pheatmap`, including with correlation-based clustering
* Understand how to define and use a helper function to repeatedly plot heatmaps for different protein subsets
* Be able to produce a scatter plot comparing log fold changes across two contrasts to identify proteins with shared or specific responses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("matrixStats")
library("tidyr")
library("dplyr")
library("tibble")
library("ggplot2")
library("ggrepel")
library("ggbeeswarm")
library("pheatmap")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


## Heatmaps

TS: INTERACTION POINTS: What are we trying to learn from a heatmap? Which proteins to plot?

A heatmap provides a two-dimensional view of protein abundances across all
samples simultaneously, and is commonly used alongside hierarchical clustering
to reveal groups of proteins and samples with similar abundance patterns.

### All replicated proteins

We start by plotting a heatmap of all proteins that were retained after
the replication filter in the statistical analysis. We rename the sample
columns using the `shortname` column from the `colData` and add a column
annotation bar to indicate group membership.

```{r, fig.height=10, fig.width=8}
quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]])

# Rename the samples with their shortname
colnames(quant_mtx) <- dia_qf$shortname

sample_annotations <- colData(dia_qf) %>%
  data.frame() %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames('shortname') %>%
  select(group)

head(sample_annotations)

pheatmap(quant_mtx,
         scale = 'row',
         show_rownames = FALSE,
         annotation_col = sample_annotations)
```

### Improving clustering with correlation-based distance

The default Euclidean distance metric used for hierarchical clustering in
`pheatmap` can be sensitive to the absolute scale of values. An alternative
is to use 1 minus the Pearson correlation as a distance metric, which focuses
on the shape of abundance profiles rather than their absolute values.

We could use `clustering_distance_rows = "correlation"` within `pheatmap`,
but this does not handle missing values. Instead, we calculate the
correlation-based distance manually using `cor` with
`use = "pairwise.complete.obs"` to handle missing values appropriately.

```{r, fig.height=10, fig.width=8}
# Calculate correlation-based distance manually, handling NAs with pairwise complete obs
dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

pheatmap(quant_mtx,
         scale = 'row',
         show_rownames = FALSE,
         annotation_col = sample_annotations,
         clustering_distance_rows = dist_rows,
         clustering_distance_cols = dist_cols)
```

### Heatmap for the most variable proteins

Plotting all proteins can make it difficult to see structure. A common
approach is to restrict the heatmap to the most variable proteins, which are
more likely to reveal biologically meaningful patterns.

```{r}
var <- rowVars(assay(dia_qf[["norm_proteins_replicated"]]), na.rm = TRUE)

most_variable <- sort(var, decreasing = TRUE)[1:50] %>% names()
```

Since we will be making multiple heatmaps with different subsets of proteins,
we define a helper function that takes a vector of UniProt IDs and produces a
heatmap. The function uses gene name and UniProt ID combined as row labels
for interpretability.

```{r, fig.height=10, fig.width=10}
plot_heatmap_for_uids <- function(uids, ...){
  quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]][uids,])
  colnames(quant_mtx) <- dia_qf$shortname

  # Use concatenation of UniprotID and Gene name as row label
  rownames(quant_mtx) <- sprintf('%s (%s)',
                                 rowData(dia_qf[["norm_proteins_replicated"]][uids,])$Genes,
                                 rownames(quant_mtx))

  dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
  dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

  pheatmap(quant_mtx,
           scale = 'row',
           annotation_col = sample_annotations,
           clustering_distance_rows = dist_rows,
           clustering_distance_cols = dist_cols,
           fontsize = 8,
           ...)
}

plot_heatmap_for_uids(most_variable)
```

TS: POSSIBLE EXERCISE POINT. PLOT THE HEATMAPS FOR THE SIGNIFICANT PROTEINS ACCORDING TO F-TEST AND PER CONTRAST T-TESTS. HOW DO THE PROTEINS AND SAMPLES CLUSTER?

### Heatmap of proteins significant by F-test

We can use our helper function to plot a heatmap restricted to proteins
with a significant overall difference across groups according to the F-test
(adjusted p-value < 0.05).

```{r, fig.height=10, fig.width=10}
# Get the significant proteins from the F-test results
sig_prot_F <- limma_results_F %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID)

plot_heatmap_for_uids(sig_prot_F)
```

### Heatmap of proteins significant in any pairwise contrast

Similarly, we can plot the proteins that are significant (adjusted p-value
< 0.05) in at least one pairwise contrast.

```{r, fig.height=10, fig.width=10}
sig_prot_contrasts <- limma_results_all_contrasts %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID) %>%
  unique()

plot_heatmap_for_uids(sig_prot_contrasts)
```

## Comparing log fold changes across contrasts

A scatter plot of log fold changes from two contrasts — here MPXV vs Control
and COVID-19 vs Control — allows us to identify proteins with shared responses
to both infections versus those with a response specific to one condition.

TS: INTERACTION POINT: Is a linear model appropriate here? I don't think so because it predicts y from x, thus assumes x is measured without error. What is a better alternative (total least squares regression). What are we trying to learn from this plot? Which proteins to label?

We first reshape the results into wide format so that each protein has a
single row with separate columns for each contrast's log fold change and
adjusted p-value. We then categorise proteins by their significance status
across the two contrasts of interest.

```{r}
# Will need to break this up into 3 stages and explain each carefully:
# 1. Reshape the limma results to have one row per protein and separate columns for each contrast's adjusted p-value and log fold change.
# 2. Create a new column 'sig' to categorize proteins based on their significance in the two contrasts we are comparing (MPXV vs Control and COVID-19 vs Control)
# 3. Plot the log fold changes against each other, coloring and labeling points based on their significance category

compare_limma_results <- limma_results_all_contrasts %>%
  select(UniprotID, Genes, adj.P.Val, logFC, contrast) %>%
  pivot_wider(names_from = contrast, values_from = c(adj.P.Val, logFC))

compare_limma_results <- compare_limma_results %>%
  mutate(sig = case_when(
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "Both",
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control >= 0.05 ~ "MPXV only",
    adj.P.Val_MPXV_control >= 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "COVID-19 only",
    TRUE ~ "None"
  ))

compare_limma_results %>%
  arrange(desc(sig)) %>%
  ggplot(aes(logFC_MPXV_control, logFC_Covid19_control)) +
  geom_point(aes(colour = sig, alpha = sig), pch = 20, size = 3) +
  geom_smooth(method = 'lm', se = FALSE, linetype = 2, colour = 'grey') +
  geom_text_repel(data = filter(compare_limma_results, sig == "Both"),
                  aes(label = Genes),
                  size = 3,
                  colour = 'grey20', max.overlaps = 100) +
  theme_minimal() +
  labs(x = "Log Fold Change (MPXV vs Control)",
       y = "Log Fold Change (COVID-19 vs Control)") +
  scale_colour_manual(values = c("Both" = "purple",
                                 "MPXV only" = "orangered4",
                                 "COVID-19 only" = "steelblue",
                                 "None" = "grey70")) +
  scale_alpha_manual(values = c("Both" = 0.8,
                                "MPXV only" = 0.8,
                                "COVID-19 only" = 0.8,
                                "None" = 0.25)) +
  theme(legend.title = element_blank())
```


::: {.callout-tip}
#### Key Points

- Heatmaps are a useful way to visualise protein abundance patterns across all samples simultaneously; restricting to significant or highly variable proteins makes them easier to interpret
- The default Euclidean distance metric in `pheatmap` can be replaced with a correlation-based distance to focus clustering on the shape of abundance profiles; this requires computing the distance matrix manually when missing values are present
- Defining a helper function for repeated plotting tasks (such as heatmaps for different protein subsets) reduces code duplication and makes the analysis easier to extend
- Scatter plots of log fold changes from two contrasts can reveal proteins with shared or condition-specific responses; reshaping results to wide format with `pivot_wider` is a convenient way to align values from different contrasts for comparison

:::

## References {-}
