---
title: Visualisation of differential abundance results
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to produce heatmaps of protein abundance data using `pheatmap`, including with correlation-based clustering
* Understand how to define and use a helper function to repeatedly plot heatmaps for different protein subsets
* Be able to produce a scatter plot comparing log fold changes across two contrasts to identify proteins with shared or specific responses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("matrixStats")
library("tidyr")
library("dplyr")
library("tibble")
library("ggplot2")
library("ggrepel")
library("pheatmap")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


Having performed differential abundance analysis, we now want to visualise the
results in ways that complement the volcano plots produced in the previous
lesson. In this lesson we produce heatmaps to display protein abundance patterns
across all samples simultaneously, and scatter plots to compare fold changes
across contrasts to identify proteins with shared or condition-specific
responses.

## Heatmaps

::: {.callout-discussion}
#### Discussion

- What are we trying to learn from a heatmap of protein abundances? What patterns would you look for?
- Which proteins would you choose to plot — all quantified proteins, significant proteins, or the most variable? What are the trade-offs of each approach?

:::

A heatmap provides a two-dimensional view of protein abundances across all
samples simultaneously, and is commonly used alongside hierarchical clustering
to reveal groups of proteins and samples with similar abundance patterns.

### All replicated proteins

We start by plotting a heatmap of all proteins that were retained after
the replication filter in the statistical analysis. We rename the sample
columns using the `shortname` column from the `colData` and add a column
annotation bar to indicate group membership.

The default Euclidean distance metric used for hierarchical clustering in
`pheatmap` can be sensitive to the absolute scale of values. An alternative
is to use 1 minus the Pearson correlation as a distance metric, which focuses
on the shape of abundance profiles rather than their absolute values.

We could use `clustering_distance_rows = "correlation"` within `pheatmap`,
but this does not handle missing values. Instead, we calculate the
correlation-based distance manually using `cor` with
`use = "pairwise.complete.obs"` to handle missing values appropriately.
We also use Ward's minimum variance method (`clustering_method = 'ward.D2'`)
for the hierarchical clustering, which tends to produce more compact,
evenly-sized clusters than the default complete linkage.

```{r, fig.height=10, fig.width=8}
quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]])

# Rename the samples with their shortname
colnames(quant_mtx) <- dia_qf$shortname

sample_annotations <- colData(dia_qf) %>%
  data.frame() %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames('shortname') %>%
  select(group)

head(sample_annotations)

# Calculate correlation-based distance manually, handling NAs with pairwise complete obs
dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

pheatmap(quant_mtx,
         scale = 'row',
         show_rownames = FALSE,
         annotation_col = sample_annotations,
         clustering_method = 'ward.D2', # Use Ward's method for hierarchical clustering
         clustering_distance_rows = dist_rows,
         clustering_distance_cols = dist_cols)
```


Since we will be making multiple heatmaps below with different subsets of proteins,
it is useful to define a helper function rather than repeating the same code
each time. This keeps each subsequent heatmap call concise and ensures all
plots are produced in the same way.

The function below takes a vector of UniProt IDs, extracts the
corresponding rows from the quantification matrix, applies the
correlation-based distance metrics, and calls `pheatmap` with consistent
formatting. 

```{r}
plot_heatmap_for_uids <- function(uids){
  quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]][uids,])
  colnames(quant_mtx) <- dia_qf$shortname

  # Use concatenation of UniprotID and Gene name as row label
  rownames(quant_mtx) <- sprintf('%s (%s)',
                                 rowData(dia_qf[["norm_proteins_replicated"]][uids,])$Genes,
                                 rownames(quant_mtx))
  # Calculate correlation-based distance manually, handling NAs with pairwise complete obs
  dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
  dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

  pheatmap(quant_mtx,
           scale = 'row',
           clustering_method = 'ward.D2', 
           annotation_col = sample_annotations,
           clustering_distance_rows = dist_rows,
           clustering_distance_cols = dist_cols,
           fontsize = 8)
}


```


### Heatmap for the most variable proteins

Plotting all proteins can make it difficult to see structure. A common
approach is to restrict the heatmap to the most variable proteins, which are
more likely to reveal biologically meaningful patterns.

```{r, fig.width=20, fig.height=10}
var <- rowVars(assay(dia_qf[["norm_proteins_replicated"]]), na.rm = TRUE)

most_variable <- sort(var, decreasing = TRUE)[1:50] %>% names()

plot_heatmap_for_uids(most_variable)
```



:::{.callout-exercise}
#### Challenge: Heatmaps of significant proteins

{{< level 2 >}}

We have already defined the `plot_heatmap_for_uids` helper function and have F-test and pairwise contrast results available. Use these to produce heatmaps for:

1. Proteins with a significant overall difference across groups according to the F-test (`adj.P.Val < 0.05`)
2. Proteins significant (`adj.P.Val < 0.05`) in at least one pairwise contrast

For each heatmap, consider:

- How do the proteins and samples cluster?
- Do the sample groups separate cleanly?
- How does restricting to significant proteins change the heatmap compared to the most variable proteins plotted above?

<details><summary>Hint</summary>Filter `limma_results_F` and `limma_results_all_contrasts` to get the UniProt IDs of significant proteins, then pass these to `plot_heatmap_for_uids`.</details>

::: {.callout-answer collapse="true"}

### Heatmap of proteins significant by F-test

```{r, fig.height=10, fig.width=10}
sig_prot_F <- limma_results_F %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID)

plot_heatmap_for_uids(sig_prot_F)
```

### Heatmap of proteins significant in any pairwise contrast

```{r, fig.height=10, fig.width=10}
sig_prot_contrasts <- limma_results_all_contrasts %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID) %>%
  unique()

plot_heatmap_for_uids(sig_prot_contrasts)
```

:::

:::

## Comparing log fold changes across contrasts

A scatter plot of log fold changes from two contrasts — here MPXV vs Control
and COVID-19 vs Control — is one way to visualise to what extent the response
to both infections is shared or infection-specific.

We first use `pivot_wider` to reshape the results so that each protein
occupies a single row, with separate columns for the log fold change and
adjusted p-value from each contrast.

```{r}
compare_limma_results <- limma_results_all_contrasts %>%
  select(UniprotID, Genes, adj.P.Val, logFC, contrast) %>%
  pivot_wider(names_from = contrast, values_from = c(adj.P.Val, logFC))
```

We can now plot the log fold changes against each other.

```{r}
compare_limma_results %>%
  ggplot(aes(x=logFC_MPXV_control, y=logFC_Covid19_control)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Log Fold Change (MPXV vs Control)",
       y = "Log Fold Change (COVID-19 vs Control)") 
```

We'd like to highlight the significance status of each protein across the two
contrasts. We can create a `sig` column that categorises each protein by its
significance status across the two contrasts, distinguishing proteins
significant in both, in only one, or in neither.

```{r}
compare_limma_results <- compare_limma_results %>%
  mutate(sig = case_when(
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "Both",
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control >= 0.05 ~ "MPXV only",
    adj.P.Val_MPXV_control >= 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "COVID-19 only",
    TRUE ~ "None"
  ))
```

Now we can re-plot the log fold changes against each other, colouring
points by significance category. We will label only those significant in
both contrasts to keep the plot legible.

```{r}
p <- compare_limma_results %>%
  arrange(desc(sig)) %>%
  ggplot(aes(x=logFC_MPXV_control, y=logFC_Covid19_control)) +
  geom_point(aes(colour = sig), pch = 20, size = 3) +
  geom_text_repel(data = filter(compare_limma_results, sig == "Both"),
                  aes(label = Genes),
                  size = 3,
                  colour = 'grey20', max.overlaps = 100) +
  theme_minimal() +
  labs(x = "Log Fold Change (MPXV vs Control)",
       y = "Log Fold Change (COVID-19 vs Control)") +
  scale_colour_manual(values = c("Both" = "purple",
                                 "MPXV only" = "orangered4",
                                 "COVID-19 only" = "steelblue",
                                 "None" = "grey70"))

print(p)
```

Finally, we map transparency (`alpha`) to the significance category to draw the eye
towards proteins significant in both contrasts, and add a linear regression line
with `geom_smooth(method = 'lm')` to show the overall relationship between log
fold changes across the two infections.

```{r}
p2 <- p + aes(alpha = sig) +
  scale_alpha_manual(values = c("Both" = 0.8,
                                "MPXV only" = 0.8,
                                "COVID-19 only" = 0.8,
                                "None" = 0.25)) +
  geom_smooth(aes(x=logFC_MPXV_control, y=logFC_Covid19_control),
              inherit.aes = FALSE, # Don't inherit aesthetics to avoid inheriting alpha aesthetics
              method = 'lm', se = FALSE, linetype = 2, colour = 'grey') +
  theme(legend.title = element_blank())

print(p2)
```

::: {.callout-discussion}
#### Discussion

- What are we trying to learn from this scatter plot — what does a protein's position in the plot tell us?
- Which proteins are most interesting?
- The plot includes a linear regression line. Is a standard linear model (Ordinary Least Squares; OLS) appropriate here, given that both axes represent quantities measured with error? What would be a better alternative?

:::


::: {.callout-tip}
#### Key Points

- Heatmaps are a useful way to visualise protein abundance patterns across all samples simultaneously; restricting to significant or highly variable proteins makes them easier to interpret
- The default Euclidean distance metric in `pheatmap` can be replaced with a correlation-based distance to focus clustering on the shape of abundance profiles; this requires computing the distance matrix manually when missing values are present
- Ward's minimum variance linkage (`clustering_method = 'ward.D2'`) tends to produce more compact, evenly-sized clusters than the default complete linkage
- Defining a helper function for repeated plotting tasks (such as heatmaps for different protein subsets) reduces code duplication and makes the analysis easier to extend
- Scatter plots of log fold changes from two contrasts can reveal proteins with shared or condition-specific responses; reshaping results to wide format with `pivot_wider` is a convenient way to align values from different contrasts for comparison

:::

## References {-}
