---
title: Visualisation of differential abundance results
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Know how to produce heatmaps of protein abundance data using `pheatmap`, including with correlation-based clustering
* Understand how to define and use a helper function to repeatedly plot heatmaps for different protein subsets
* Be able to produce a scatter plot comparing log fold changes across two contrasts to identify proteins with shared or specific responses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("matrixStats")
library("tidyr")
library("dplyr")
library("tibble")
library("ggplot2")
library("ggrepel")
library("ggbeeswarm")
library("pheatmap")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


## Heatmaps

::: {.callout-discussion}
#### Discussion

- What are we trying to learn from a heatmap of protein abundances? What patterns would you look for?
- Which proteins would you choose to plot — all quantified proteins, significant proteins, or the most variable? What are the trade-offs of each approach?

:::

A heatmap provides a two-dimensional view of protein abundances across all
samples simultaneously, and is commonly used alongside hierarchical clustering
to reveal groups of proteins and samples with similar abundance patterns.

### All replicated proteins

We start by plotting a heatmap of all proteins that were retained after
the replication filter in the statistical analysis. We rename the sample
columns using the `shortname` column from the `colData` and add a column
annotation bar to indicate group membership.

```{r, fig.height=10, fig.width=8}
quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]])

# Rename the samples with their shortname
colnames(quant_mtx) <- dia_qf$shortname

sample_annotations <- colData(dia_qf) %>%
  data.frame() %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames('shortname') %>%
  select(group)

head(sample_annotations)

pheatmap(quant_mtx,
         scale = 'row',
         show_rownames = FALSE,
         annotation_col = sample_annotations)
```

### Improving clustering with correlation-based distance

The default Euclidean distance metric used for hierarchical clustering in
`pheatmap` can be sensitive to the absolute scale of values. An alternative
is to use 1 minus the Pearson correlation as a distance metric, which focuses
on the shape of abundance profiles rather than their absolute values.

We could use `clustering_distance_rows = "correlation"` within `pheatmap`,
but this does not handle missing values. Instead, we calculate the
correlation-based distance manually using `cor` with
`use = "pairwise.complete.obs"` to handle missing values appropriately.

```{r, fig.height=10, fig.width=8}
# Calculate correlation-based distance manually, handling NAs with pairwise complete obs
dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

pheatmap(quant_mtx,
         scale = 'row',
         show_rownames = FALSE,
         annotation_col = sample_annotations,
         clustering_distance_rows = dist_rows,
         clustering_distance_cols = dist_cols)
```

### Heatmap for the most variable proteins

Plotting all proteins can make it difficult to see structure. A common
approach is to restrict the heatmap to the most variable proteins, which are
more likely to reveal biologically meaningful patterns.

```{r}
var <- rowVars(assay(dia_qf[["norm_proteins_replicated"]]), na.rm = TRUE)

most_variable <- sort(var, decreasing = TRUE)[1:50] %>% names()
```

Since we will be making multiple heatmaps with different subsets of proteins,
we define a helper function that takes a vector of UniProt IDs and produces a
heatmap. The function uses gene name and UniProt ID combined as row labels
for interpretability.

```{r, fig.height=10, fig.width=10}
plot_heatmap_for_uids <- function(uids, ...){
  quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]][uids,])
  colnames(quant_mtx) <- dia_qf$shortname

  # Use concatenation of UniprotID and Gene name as row label
  rownames(quant_mtx) <- sprintf('%s (%s)',
                                 rowData(dia_qf[["norm_proteins_replicated"]][uids,])$Genes,
                                 rownames(quant_mtx))

  dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
  dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

  pheatmap(quant_mtx,
           scale = 'row',
           annotation_col = sample_annotations,
           clustering_distance_rows = dist_rows,
           clustering_distance_cols = dist_cols,
           fontsize = 8,
           ...)
}

plot_heatmap_for_uids(most_variable)
```

:::{.callout-exercise}
#### Challenge: Heatmaps of significant proteins

{{< level 2 >}}

We have already defined the `plot_heatmap_for_uids` helper function and have F-test and pairwise contrast results available. Use these to produce heatmaps for:

1. Proteins with a significant overall difference across groups according to the F-test (`adj.P.Val < 0.05`)
2. Proteins significant (`adj.P.Val < 0.05`) in at least one pairwise contrast

For each heatmap, consider:

- How do the proteins and samples cluster?
- Do the sample groups separate cleanly?
- How does restricting to significant proteins change the heatmap compared to the most variable proteins plotted above?

<details><summary>Hint</summary>Filter `limma_results_F` and `limma_results_all_contrasts` to get the UniProt IDs of significant proteins, then pass these to `plot_heatmap_for_uids`.</details>

::: {.callout-answer collapse="true"}

### Heatmap of proteins significant by F-test

```{r, fig.height=10, fig.width=10}
sig_prot_F <- limma_results_F %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID)

plot_heatmap_for_uids(sig_prot_F)
```

### Heatmap of proteins significant in any pairwise contrast

```{r, fig.height=10, fig.width=10}
sig_prot_contrasts <- limma_results_all_contrasts %>%
  filter(adj.P.Val < 0.05) %>%
  pull(UniprotID) %>%
  unique()

plot_heatmap_for_uids(sig_prot_contrasts)
```

:::

:::

## Comparing log fold changes across contrasts

A scatter plot of log fold changes from two contrasts — here MPXV vs Control
and COVID-19 vs Control — allows us to identify proteins with shared responses
to both infections versus those with a response specific to one condition.

::: {.callout-discussion}
#### Discussion

- What are we trying to learn from this scatter plot — what does a protein's position in the plot tell us?
- The plot includes a linear regression line. Is a standard linear model (OLS) appropriate here, given that both axes represent quantities measured with error? What would be a better alternative?
- Which proteins would you choose to label, and why?

:::

We first reshape the results into wide format so that each protein has a
single row with separate columns for each contrast's log fold change and
adjusted p-value. We then categorise proteins by their significance status
across the two contrasts of interest.

```{r}
# Will need to break this up into 3 stages and explain each carefully:
# 1. Reshape the limma results to have one row per protein and separate columns for each contrast's adjusted p-value and log fold change.
# 2. Create a new column 'sig' to categorize proteins based on their significance in the two contrasts we are comparing (MPXV vs Control and COVID-19 vs Control)
# 3. Plot the log fold changes against each other, coloring and labeling points based on their significance category

compare_limma_results <- limma_results_all_contrasts %>%
  select(UniprotID, Genes, adj.P.Val, logFC, contrast) %>%
  pivot_wider(names_from = contrast, values_from = c(adj.P.Val, logFC))

compare_limma_results <- compare_limma_results %>%
  mutate(sig = case_when(
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "Both",
    adj.P.Val_MPXV_control < 0.05 & adj.P.Val_Covid19_control >= 0.05 ~ "MPXV only",
    adj.P.Val_MPXV_control >= 0.05 & adj.P.Val_Covid19_control < 0.05 ~ "COVID-19 only",
    TRUE ~ "None"
  ))

compare_limma_results %>%
  arrange(desc(sig)) %>%
  ggplot(aes(logFC_MPXV_control, logFC_Covid19_control)) +
  geom_point(aes(colour = sig, alpha = sig), pch = 20, size = 3) +
  geom_smooth(method = 'lm', se = FALSE, linetype = 2, colour = 'grey') +
  geom_text_repel(data = filter(compare_limma_results, sig == "Both"),
                  aes(label = Genes),
                  size = 3,
                  colour = 'grey20', max.overlaps = 100) +
  theme_minimal() +
  labs(x = "Log Fold Change (MPXV vs Control)",
       y = "Log Fold Change (COVID-19 vs Control)") +
  scale_colour_manual(values = c("Both" = "purple",
                                 "MPXV only" = "orangered4",
                                 "COVID-19 only" = "steelblue",
                                 "None" = "grey70")) +
  scale_alpha_manual(values = c("Both" = 0.8,
                                "MPXV only" = 0.8,
                                "COVID-19 only" = 0.8,
                                "None" = 0.25)) +
  theme(legend.title = element_blank())
```


::: {.callout-tip}
#### Key Points

- Heatmaps are a useful way to visualise protein abundance patterns across all samples simultaneously; restricting to significant or highly variable proteins makes them easier to interpret
- The default Euclidean distance metric in `pheatmap` can be replaced with a correlation-based distance to focus clustering on the shape of abundance profiles; this requires computing the distance matrix manually when missing values are present
- Defining a helper function for repeated plotting tasks (such as heatmaps for different protein subsets) reduces code duplication and makes the analysis easier to extend
- Scatter plots of log fold changes from two contrasts can reveal proteins with shared or condition-specific responses; reshaping results to wide format with `pivot_wider` is a convenient way to align values from different contrasts for comparison

:::

## References {-}
