---
title: Functional enrichment analysis
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Complete GO over-representation analysis using the `enrichGO` function from the `clusterProfiler` package
* Know how to reduce redundancy in GO enrichment results using semantic similarity with the `GOSemSim` package
* Be able to visualise and interpret the results of functional enrichment analyses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("ggplot2")
library("clusterProfiler")
library("tidyr")
library("tibble")
library("org.Hs.eg.db")
library("enrichplot")
library("GOSemSim")
library("pheatmap")
library("dplyr")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


## Biological interpretation of differentially abundant proteins

Having identified proteins with significantly different abundances between
groups, we now aim to understand the biological processes and pathways these
proteins are associated with. We use **Gene Ontology (GO) over-representation
analysis (ORA)**, which tests whether a selected list of proteins (e.g., those
significantly decreased in COVID-19 vs Control) is enriched for particular GO
terms relative to a background set of all quantified proteins.

We focus our analysis on the COVID-19 vs Control contrast, as this is the
comparison of primary biological interest.

```{r}
limma_results_covid_control <- limma_results_all_contrasts %>%
  filter(contrast == 'Covid19_control')
```

## GO over-representation analysis

{{< include ../_shared/_go_enrichment_theory.md >}}

TS: I wouldn't ever use `clusterProfiler` in my own work to perform GO term ORA as I'd want to use the GO annotations direct to the UniprotIDs. Also, where mapping is required, I'd want to have control/insight in the mapping of my protein IDs to gene IDs and the GO annotations. I would also want to explore whether there's a relationship between abundance and probability of being significant as this is likely to be an artifact that needs to be accounted for using, e.g `goseq` package. However, `clusterProfiler` is a popular package and is easy to use.

We use the `enrichGO` function from the `clusterProfiler` package to perform
GO ORA on the proteins with significantly decreased abundance in COVID-19 patients
relative to controls (adjusted p-value < 0.05 and negative log fold change).
We test all three GO ontologies simultaneously (`ont = "ALL"`).

```{r}
sig_down <- limma_results_covid_control %>%
  filter(adj.P.Val < 0.05, logFC < 0) %>%
  pull(UniprotID)

go_down <- enrichGO(gene = sig_down,                                # list of down proteins
                    universe = limma_results_covid_control$UniprotID, # all proteins
                    OrgDb = org.Hs.eg.db,                           # database to query
                    keyType = "UNIPROT",                            # protein ID encoding
                    pvalueCutoff = 0.05,
                    ont = "ALL",                                    # can be CC, MF, BP, or ALL
                    readable = TRUE)
```

### Visualising significant GO terms

::: {.callout-discussion}
#### Discussion

- Looking at the GO enrichment results, why do we see so many similar or overlapping terms?
- What property of the Gene Ontology leads to this redundancy, and how might we address it?

:::

We produce a dotplot of the significant GO terms, faceted by ontology
(Biological Process, Molecular Function, Cellular Component). The size of
each dot represents the number of proteins annotated with that term, and
the colour indicates the adjusted p-value.

```{r}

dotplot(go_down,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Reducing GO term redundancy using semantic similarity

GO terms are arranged in a hierarchy, meaning many significant terms will
be related to one another. This leads to a long list of redundant results.
The `GOSemSim` package can be used to compute pairwise semantic similarity
between terms, and the `simplify` function from `clusterProfiler` can then
remove highly similar terms, keeping only the most significant representative
from each cluster.

```{r}

gd <- godata('org.Hs.eg.db', ont = "BP")

go_down_similarity <- pairwise_termsim(go_down, method = "Wang", semData = gd)

go_down_simplified <- simplify(go_down_similarity,
                               cutoff = 0.7, # similarity threshold for clustering
                               by = "p.adjust",
                               select_fun = min)
```

We can now re-visualise the simplified results, which should be more interpretable with less redundancy.

```{r}
dotplot(go_down_simplified,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Visualising relationships between GO terms with a treeplot

The `treeplot` function from `enrichplot` hierarchically clusters the
significant GO terms based on their pairwise semantic similarity and
displays them as a tree, making the relationships between terms easier
to interpret.


```{r, fig.height=6, fig.width=10}
# TS: treeplot has some fairly horrible arguments to control the plot parameters
# and stop labels overlapping. Honestly, I just played with the values until it
# worked!
  
treeplot(go_down_similarity,
         showCategory = Inf,
         nCluster = 3,
         fontsize_tiplab = 2,
         fontsize_cladelab = 3,
         cladelab_offset = 6,
         tiplab_offset = 0.5,
         hexpand = 0.2)
```

The treeplot provides a complementary view to the dotplot — rather than simply
listing enriched terms, it reveals their biological relationships. Terms that
cluster together share semantic similarity and likely reflect the same
underlying biology. This makes it easier to summarise the enrichment results
in terms of a small number of coherent biological themes rather than a long
list of individually significant terms.


## Linking GO results back to protein abundances

The GO enrichment results tell us which biological processes are over-represented
among our proteins of interest, but it is often informative to go back and
examine the sample-level abundance patterns for the proteins driving that
enrichment.

Because we set `readable = TRUE` in our `enrichGO` call, the `geneID` column of
the results contains gene symbols rather than Entrez IDs. We can extract
these directly from the enrichment result and use `AnnotationDbi::select` to map
them to UniProt accessions, filtering to proteins present in our experiment.


```{r}

pm_lipo_part_clear <- go_down@result %>%
  filter(Description == "plasma lipoprotein particle clearance") %>%
  pull(geneID) %>%
  strsplit("/") %>%
  unlist() 

# Gene symbol -> UniProt map
g2p <- AnnotationDbi::select(org.Hs.eg.db, keys = pm_lipo_part_clear,
                             columns = "UNIPROT", keytype = "SYMBOL") %>%
  filter(UNIPROT %in% limma_results_covid_control$UniprotID)

print(g2p)
```

:::{.callout-exercise}
#### Challenge: Visualising abundances of plasma lipoprotein particle clearance proteins

{{< level 3 >}}

Use the UniProt IDs retrieved above to assess whether the protein abundances for the 'plasma lipoprotein particle clearance' proteins are reliable. Then examine the abundance of these proteins across the samples using a heatmap. Do the samples cluster by condition on these proteins alone? 

1. Use `subsetByFeature` to subset `dia_qf` to the features belonging to these
   proteins. Then plot the precursor-level and protein-level abundances side by
   side across samples, adapting the approach from the protein exploration
   notebook.
2. Plot a heatmap of the protein-level abundances for these proteins, with
   samples annotated by group. Do the samples separate as you would expect from
   the enrichment result?
   

<details><summary>Hint</summary>Pass `rowvars = 'Genes'` to `longForm` to carry gene names through to the plot, then facet by gene name and assay.</details>

::: {.callout-answer collapse="true"}

```{r, fig.height=10, fig.width=20}
# extract UniProt IDs for the proteins driving the enrichment of the 'plasma lipoprotein particle clearance' term
uids <- g2p$UNIPROT


# 1. Subset to the proteins of interest and
# plot precursor and protein level abundances across samples
dia_qf_lipids <- subsetByFeature(dia_qf, uids)

dia_qf_lipids[, , c("precursors_filtered_missing", "proteins")] %>%
  longForm(rowvars=c('Genes')) %>%
  as_tibble() %>%
  mutate(assay_order = factor(assay,
                              levels = c("precursors_filtered_missing",
                                         "proteins"))) %>%
  mutate(value=ifelse(assay=='precursors_filtered_missing', log2(value), value)) %>%
  ggplot(aes(x = colname, y = value, colour = assay)) +
  geom_point(size=0.5) +
  geom_line(aes(group = rowname)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size = 7)) +
  facet_grid(Genes~assay_order,scales='free') +
  labs(x='Sample', y='Abundance (log2)') +
  theme(aspect.ratio=1/2)
```

```{r, fig.height=6, fig.width=6}

# 2. Protein-level heatmap for plasma lipoprotein particle clearance proteins
sample_annotations <- colData(dia_qf) %>%
  data.frame() %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames("shortname") %>%
  select(group)

quant_mtx <- assay(dia_qf[["norm_proteins_replicated"]][uids, ])
colnames(quant_mtx) <- dia_qf$shortname

rownames(quant_mtx) <- rowData(dia_qf[["norm_proteins_replicated"]][uids, ])$Genes

dist_cols <- as.dist(1 - cor(quant_mtx, use = "pairwise.complete.obs"))
dist_rows <- as.dist(1 - cor(t(quant_mtx), use = "pairwise.complete.obs"))

pheatmap(quant_mtx,
         scale = "row",
         annotation_col = sample_annotations,
         clustering_distance_rows = dist_rows,
         clustering_distance_cols = dist_cols,
         clustering_method = 'ward.D2',
         fontsize = 8)
```



:::

:::

::: {.callout-tip}
#### Key Points


- GO over-representation analysis tests whether a set of proteins of interest is enriched for particular biological annotations relative to a background set; the background should be all proteins quantified in the experiment
- GO results often contain many redundant terms due to the hierarchical nature of the GO; semantic similarity-based methods (e.g., `GOSemSim`) can reduce this redundancy to make results more interpretable
- Visualising the relationships between significant GO terms (e.g., with `enrichplot::treeplot`) can help to summarise results in terms of a small number of coherent biological themes rather than a long list of individual terms
- Linking GO results back to the underlying data can help to validate and interpret the enrichment results, and reveal whether the proteins driving the enrichment show the expected abundance patterns across samples
:::

## References {-}
