---
title: Functional enrichment analysis
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Complete GO over-representation analysis using the `enrichGO` function from the `clusterProfiler` package
* Know how to reduce redundancy in GO enrichment results using semantic similarity with the `GOSemSim` package
* Understand how to perform Gene Set Enrichment Analysis (GSEA) using ranked protein fold changes with the `fgsea` package
* Be able to visualise and interpret the results of functional enrichment analyses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("ggplot2")
library("ggrepel")
library("fgsea")
library("clusterProfiler")
library("tidyr")
library("dplyr")
library("tibble")
library("org.Hs.eg.db")
library("enrichplot")
library("GOSemSim")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


## Biological interpretation of differentially abundant proteins

Having identified proteins with significantly different abundances between
groups, we now aim to understand the biological processes and pathways these
proteins are associated with. Two complementary approaches are used here:

1. **Gene Ontology (GO) over-representation analysis (ORA)** — tests whether
   a selected list of proteins (e.g., those significantly decreased in MPXV
   vs Control) is enriched for particular GO terms relative to a background
   set of all quantified proteins.

2. **Gene Set Enrichment Analysis (GSEA)** — uses a ranked list of all
   proteins (ranked by a metric combining fold change and significance) to
   identify pathways that are coordinately up- or down-regulated, without
   requiring a hard significance threshold.

TS: I think we should focus on one and go into more depth on that, rather than trying to cover both ORA and GSEA. ORA is more commonly used and easier to understand, so I suggest we focus on that. We can briefly mention GSEA as an alternative approach, but I don't think we have time to go through it in detail. I haven't added exercises or discussion points for the GSEA section, but we could if we want to cover it.

We focus our analysis on the MPXV vs Control contrast, as this is the
comparison of primary biological interest.

```{r}
limma_results_mpxv_control <- limma_results_all_contrasts %>%
  filter(contrast == 'MPXV_control')
```

## GO over-representation analysis

{{< include ../_shared/_go_enrichment_theory.md >}}

TS: I wouldn't ever use `clusterProfiler` in my own work to perform GO term ORA as I'd want to use the GO annotations direct to the UniprotIDs. Also, where mapping is required, I'd want to have control/insight in the mapping of my protein IDs to gene IDs and the GO annotations. I would also want to explore whether there's a relationship between abundance and probability of being significant as this is likely to be an artifact that needs to be accounted for using, e.g `goseq` package. However, `clusterProfiler` is a popular package and is easy to use.

We use the `enrichGO` function from the `clusterProfiler` package to perform
GO ORA on the proteins with significantly decreased abundance in MPXV patients
relative to controls (adjusted p-value < 0.05 and negative log fold change).
We test all three GO ontologies simultaneously (`ont = "ALL"`).

```{r}
sig_down <- limma_results_mpxv_control %>%
  filter(adj.P.Val < 0.05, logFC < 0) %>%
  pull(UniprotID)

go_down <- enrichGO(gene = sig_down,                                # list of down proteins
                    universe = limma_results_mpxv_control$UniprotID, # all proteins
                    OrgDb = org.Hs.eg.db,                           # database to query
                    keyType = "UNIPROT",                            # protein ID encoding
                    qvalueCutoff = 0.05,
                    ont = "ALL",                                    # can be CC, MF, BP, or ALL
                    readable = TRUE)
```

### Visualising significant GO terms

::: {.callout-discussion}
#### Discussion

- Looking at the GO enrichment results, why do we see so many similar or overlapping terms?
- What property of the Gene Ontology leads to this redundancy, and how might we address it?

:::

We produce a dotplot of the significant GO terms, faceted by ontology
(Biological Process, Molecular Function, Cellular Component). The size of
each dot represents the number of proteins annotated with that term, and
the colour indicates the adjusted p-value.

```{r}
# Filter to the significant GO terms
sig_go_down <- go_down %>%
  filter(p.adjust < 0.05)

dotplot(sig_go_down,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Reducing GO term redundancy using semantic similarity

GO terms are arranged in a hierarchy, meaning many significant terms will
be related to one another. This leads to a long list of redundant results.
The `GOSemSim` package can be used to compute pairwise semantic similarity
between terms, and the `simplify` function from `clusterProfiler` can then
remove highly similar terms, keeping only the most significant representative
from each cluster.

```{r}
d <- godata('org.Hs.eg.db', ont = "BP")

sig_go_down_similarity <- pairwise_termsim(sig_go_down, method = "Wang", semData = d)

sig_go_down_simplified <- simplify(sig_go_down_similarity,
                                   cutoff = 0.7,
                                   by = "p.adjust",
                                   select_fun = min)
```

We can now re-visualise the simplified results, which should be more interpretable with less redundancy.

```{r}
dotplot(sig_go_down_simplified,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Visualising relationships between GO terms with a treeplot

The `treeplot` function from `enrichplot` hierarchically clusters the
significant GO terms based on their pairwise semantic similarity and
displays them as a tree, making the relationships between terms easier
to interpret.


```{r, fig.height=6, fig.width=10}
# TS: treeplot has some fairly horrible arguments to control the plot parameters
# and stop labels overlapping. Honestly, I just played with the values until it
# worked!
  
treeplot(sig_go_down_similarity,
         showCategory = Inf,
         fontsize_tiplab = 2,
         fontsize_cladelab = 3,
         cladelab_offset = 6,
         tiplab_offset = 0.5,
         hexpand = 0.2)
```

## Reactome Gene Set Enrichment Analysis

TS: We can't use clusterProfiler for GSEA unless we first manually map from UniprotID to Entrez. Given that REACTOME annotations are available for UniprotIDs, I favour using these and the `fgsea` package for GSEA. The code to obtain the REACTOME annotations is available in the extended materials, but we won't go through it in detail here. The key point is that we need to expand the annotations to include all parent terms, and then map from Reactome IDs to pathway names for interpretability. The resulting objects are a list of pathways with their associated UniprotIDs (`reactome_annot_list`) and a mapping from Reactome IDs to pathway names (`reactome_map`).

ORA requires a binary classification of proteins into significant and
non-significant groups, which can be sensitive to the choice of threshold.
GSEA instead uses a ranking of all proteins to identify pathways that are
coordinately shifted in abundance, making it complementary to ORA.

Since Reactome pathway annotations are available for UniProt IDs directly,
we use these rather than requiring a mapping to Entrez gene IDs. The code
to obtain and process the Reactome annotations is provided in the
[extended materials](extended/09_reactome_annotations.html).
Here we load the pre-computed objects.

```{r}
reactome_annot_list <- readRDS(here("course_files/dia/preprocessed/reactome_annot_list.rds"))
reactome_map        <- readRDS(here("course_files/dia/preprocessed/reactome_map.rds"))

# Examine the objects
head(reactome_annot_list)
head(reactome_map)

```


### Running GSEA with `fgsea`

We rank all proteins by a metric combining log fold change and significance:
log fold change multiplied by -log10(p-value). This gives a single ranking
value that reflects both the magnitude and the confidence of the change.
Proteins with large positive values are strongly and significantly increased;
large negative values are strongly and significantly decreased.

```{r}
# Make fgsea results deterministic
set.seed(0)

stats <- limma_results_mpxv_control$logFC * -log10(limma_results_mpxv_control$P.Value)
names(stats) <- limma_results_mpxv_control$UniprotID

# run fgsea with minSize and maxSize set to filter out very small and very large pathways
gsea_reactome <- fgsea(pathways = reactome_annot_list,
                       stats = stats,
                       minSize = 3,
                       maxSize = 300,
                       nPermSimple = 100000) %>%
  dplyr::rename(Reactome = pathway) %>%
  filter(lengths(leadingEdge) > 2) # filter out pathways with very small leading edges
```

We merge pathway names for interpretability and arrange by p-value.

```{r}
gsea_reactome <- gsea_reactome %>%
  merge(reactome_map, by = 'Reactome') %>%
  arrange(pval)
```

### Visualising GSEA results

We plot the Normalised Enrichment Score (NES) against -log10(adjusted
p-value) to show both the strength of enrichment and its significance,
colouring points by NES to distinguish positively and negatively enriched
pathways. The most significant pathways are labelled.

```{r, fig.width=8, fig.height=6}
gsea_reactome_highlight <- gsea_reactome %>%
  filter(padj < 0.1) %>%
  group_by(sign(NES)) %>%
  slice_min(order_by = pval, n = 5)

gsea_reactome %>%
  ggplot(aes(NES, -log10(padj), colour = NES, alpha = abs(NES))) +
  geom_point() +
  labs(x = "Normalized Enrichment Score (NES)", y = "-log10(Adjusted P-value)") +
  theme_minimal() +
  ggrepel::geom_text_repel(data = gsea_reactome_highlight,
                           aes(label = Pathway),
                           size = 2.5,
                           colour = 'grey20',
                           max.overlaps = 100,
                           force = 10) +
  theme(legend.position = "none") +
  scale_colour_gradient2(low = 'steelblue', high = 'orangered4',
                         mid = 'white', midpoint = 0)
```


::: {.callout-tip}
#### Key Points

- GO over-representation analysis tests whether a set of proteins of interest is enriched for particular biological annotations relative to a background set; the background should be all proteins quantified in the experiment
- GO results often contain many redundant terms due to the hierarchical nature of the GO; semantic similarity-based methods (e.g., `GOSemSim`) can reduce this redundancy to make results more interpretable
- GSEA uses a ranked list of all proteins rather than a binary significant/non-significant split, making it complementary to ORA and less sensitive to arbitrary significance thresholds
- Reactome pathway annotations are available at the UniProt ID level, enabling GSEA without the need to map to gene IDs; the `fgsea` package provides a fast implementation of GSEA
- The normalised enrichment score (NES) from GSEA reflects both the direction and strength of pathway enrichment; combining this with the adjusted p-value in a single plot helps prioritise pathways for further investigation

:::

## References {-}
