---
title: Functional enrichment analysis
bibliography: ../course_refs.bib
editor_options:
  chunk_output_type: console
---

::: {.callout-tip}
#### Learning Objectives

* Complete GO over-representation analysis using the `enrichGO` function from the `clusterProfiler` package
* Know how to reduce redundancy in GO enrichment results using semantic similarity with the `GOSemSim` package
* Be able to visualise and interpret the results of functional enrichment analyses

:::


```{r, eval=TRUE, include=FALSE}
library("here")
library("QFeatures")
library("ggplot2")
library("clusterProfiler")
library("tidyr")
library("dplyr")
library("tibble")
library("org.Hs.eg.db")
library("enrichplot")
library("GOSemSim")

load(here("course_files/dia/preprocessed/lesson06.rda"), verbose = TRUE)
```


## Biological interpretation of differentially abundant proteins

Having identified proteins with significantly different abundances between
groups, we now aim to understand the biological processes and pathways these
proteins are associated with. We use **Gene Ontology (GO) over-representation
analysis (ORA)**, which tests whether a selected list of proteins (e.g., those
significantly decreased in MPXV vs Control) is enriched for particular GO
terms relative to a background set of all quantified proteins.

We focus our analysis on the MPXV vs Control contrast, as this is the
comparison of primary biological interest.

```{r}
limma_results_mpxv_control <- limma_results_all_contrasts %>%
  filter(contrast == 'MPXV_control')
```

## GO over-representation analysis

{{< include ../_shared/_go_enrichment_theory.md >}}

TS: I wouldn't ever use `clusterProfiler` in my own work to perform GO term ORA as I'd want to use the GO annotations direct to the UniprotIDs. Also, where mapping is required, I'd want to have control/insight in the mapping of my protein IDs to gene IDs and the GO annotations. I would also want to explore whether there's a relationship between abundance and probability of being significant as this is likely to be an artifact that needs to be accounted for using, e.g `goseq` package. However, `clusterProfiler` is a popular package and is easy to use.

We use the `enrichGO` function from the `clusterProfiler` package to perform
GO ORA on the proteins with significantly decreased abundance in MPXV patients
relative to controls (adjusted p-value < 0.05 and negative log fold change).
We test all three GO ontologies simultaneously (`ont = "ALL"`).

```{r}
sig_down <- limma_results_mpxv_control %>%
  filter(adj.P.Val < 0.05, logFC < 0) %>%
  pull(UniprotID)

go_down <- enrichGO(gene = sig_down,                                # list of down proteins
                    universe = limma_results_mpxv_control$UniprotID, # all proteins
                    OrgDb = org.Hs.eg.db,                           # database to query
                    keyType = "UNIPROT",                            # protein ID encoding
                    pvalueCutoff = 0.05,
                    ont = "ALL",                                    # can be CC, MF, BP, or ALL
                    readable = TRUE)
```

### Visualising significant GO terms

::: {.callout-discussion}
#### Discussion

- Looking at the GO enrichment results, why do we see so many similar or overlapping terms?
- What property of the Gene Ontology leads to this redundancy, and how might we address it?

:::

We produce a dotplot of the significant GO terms, faceted by ontology
(Biological Process, Molecular Function, Cellular Component). The size of
each dot represents the number of proteins annotated with that term, and
the colour indicates the adjusted p-value.

```{r}
# Filter to the significant GO terms
sig_go_down <- go_down %>%
  filter(p.adjust < 0.05)

dotplot(sig_go_down,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Reducing GO term redundancy using semantic similarity

GO terms are arranged in a hierarchy, meaning many significant terms will
be related to one another. This leads to a long list of redundant results.
The `GOSemSim` package can be used to compute pairwise semantic similarity
between terms, and the `simplify` function from `clusterProfiler` can then
remove highly similar terms, keeping only the most significant representative
from each cluster.

```{r}
d <- godata('org.Hs.eg.db', ont = "BP")

sig_go_down_similarity <- pairwise_termsim(sig_go_down, method = "Wang", semData = d)

sig_go_down_simplified <- simplify(sig_go_down_similarity,
                                   cutoff = 0.7,
                                   by = "p.adjust",
                                   select_fun = min)
```

We can now re-visualise the simplified results, which should be more interpretable with less redundancy.

```{r}
dotplot(sig_go_down_simplified,
        x = "Count",
        font.size = 10,
        showCategory = Inf,
        label_format = 100,
        split = "ONTOLOGY",
        color = "p.adjust") +
  facet_grid(ONTOLOGY~., scales = 'free_y', space = 'free_y')
```

### Visualising relationships between GO terms with a treeplot

The `treeplot` function from `enrichplot` hierarchically clusters the
significant GO terms based on their pairwise semantic similarity and
displays them as a tree, making the relationships between terms easier
to interpret.


```{r, fig.height=6, fig.width=10}
# TS: treeplot has some fairly horrible arguments to control the plot parameters
# and stop labels overlapping. Honestly, I just played with the values until it
# worked!
  
treeplot(sig_go_down_similarity,
         showCategory = Inf,
         nCluster = 3,
         fontsize_tiplab = 2,
         fontsize_cladelab = 3,
         cladelab_offset = 6,
         tiplab_offset = 0.5,
         hexpand = 0.2)
```

The treeplot provides a complementary view to the dotplot — rather than simply
listing enriched terms, it reveals their biological relationships. Terms that
cluster together share semantic similarity and likely reflect the same
underlying biology. This makes it easier to summarise the enrichment results
in terms of a small number of coherent biological themes rather than a long
list of individually significant terms.

:::{.callout-exercise}
#### Challenge: GO enrichment of proteins increased in MPXV

{{< level 2 >}}

We have focused so far on proteins with significantly *decreased* abundance in
MPXV patients relative to controls. Now apply the same GO over-representation
analysis to proteins with significantly *increased* abundance.

1. Filter `limma_results_mpxv_control` to proteins with `adj.P.Val < 0.05` and
   positive `logFC` to define the gene list.
2. Run `enrichGO` using the same settings as above.
3. What do you conclude from the results? Are there any interesting biological themes emerging from the proteins with increased abundance in MPXV patients?

::: {.callout-answer collapse="true"}

```{r}
sig_up <- limma_results_mpxv_control %>%
  filter(adj.P.Val < 0.05, logFC > 0) %>%
  pull(UniprotID)

go_up <- enrichGO(gene = sig_up,
                  universe = limma_results_mpxv_control$UniprotID,
                  OrgDb = org.Hs.eg.db,
                  keyType = "UNIPROT",
                  pvalueCutoff = 0.05,
                  ont = "ALL",
                  readable = TRUE)

# There are no over-represented GO terms!
print(go_up)
```

:::

:::

::: {.callout-tip}
#### Key Points

- GO over-representation analysis tests whether a set of proteins of interest is enriched for particular biological annotations relative to a background set; the background should be all proteins quantified in the experiment
- GO results often contain many redundant terms due to the hierarchical nature of the GO; semantic similarity-based methods (e.g., `GOSemSim`) can reduce this redundancy to make results more interpretable

:::

## References {-}
