---
title: Biological interpretation
bibliography: course_refs.bib
---

::: {.callout-tip}
#### Learning Objectives

* Be aware of different analyses that can be done to gain a biological understanding of expression proteomics results
* Understand the concept of Gene Ontology (GO) enrichment analyses 
* Complete GO enrichment analyses using the `enrichGO` function from the `clusterProfiler` package

:::


```{r, eval=TRUE, include=FALSE}
library("QFeatures")
library("ggplot2")
library("stringr")
library("dplyr")
library("tibble")
library("NormalyzerDE")
library("corrplot")
library("Biostrings")
library("limma")
library("statmod")
library("org.Hs.eg.db")
library("clusterProfiler")
library("enrichplot")
load("output/lesson05.rda", verbose = TRUE)
```

## Subset differentially abundant proteins

Before we can look any further into the biological meaning of any protein 
abundance changes we need to extract these proteins from our overall results.
It is also useful to re-add information about the master protein descriptions
since this is lost in the output of `limma` analysis.

```{r}
## Add master protein descriptions back
M_Desynch_results$protein_description <- all_proteins %>%
  rowData() %>%
  as_tibble() %>%
  pull(Master.Protein.Descriptions)

## Extract proteins that have significant abundance changes in M vs. Desynch
sig_changing <- M_Desynch_results %>% 
  as_tibble() %>%
  filter(significance == "sig")

sig_up <- sig_changing %>%
  filter(direction == "up")

sig_down <- sig_changing %>%
  filter(direction == "down")
```


## Biological interpretation of differentially abundant proteins

Our statistical analyses provided us with a list of proteins that are present
with significantly different abundances between different stages of the cell 
cycle. We can get an initial idea about what these proteins are and do by 
looking at the protein descriptions. 

```{r}
## Look at descriptions of proteins upregulated in M relative to deysnchronized cells
sig_up %>%
  select(protein_description) %>%
  head()
```

Whilst we may recognise some of the changing proteins, this might be the first
time that we are coming across others. Moreover, some protein descriptions 
contain useful information, but this is very limited. We still want to find out
more about the biological role of the statistically significant proteins so that
we can infer the potential effects of their abundance changes. 

There are many functional analyses that could be done on the proteins with
differential abundance:

* Investigate the biological pathways that the proteins function within (KEGG etc.)
* Identify potential interacting partners (IntAct, String.dc etc.)
* Determine the subcellular localisation in which the changing proteins are found 
* Understand the co-regulation of their mRNAs (Expression Atlas)
* Compare our changing proteins to those previously identified in other proteomic
studies of the cell cycle

### Gene Ontology (GO) enrichment analysis 

One of the common methods used to probe the biological relevance of proteins with
significant changes in abundance between conditions is to carry out Gene Ontology 
(GO) enrichment, or over-representation, analysis. 

The Gene Ontology consortium have defined a set of hierarchical descriptions to 
be assigned to genes and their resulting proteins. These descriptions are split
into three categories: cellular components (CC), biological processes (BP) and
molecular function (MF). The idea is to provide information about a protein's
subcellular localisation, functionality and which processes it contributes to
within the cell. Hence, the overarching aim of GO enrichment analysis is to 
answer the question: 

*"Given a list of proteins found to be differentially abundant in my phenotype of interest, what are the cellular components, molecular functions and biological processes involved in this phenotype?".*

Unfortunately, just looking at the GO terms associated with our differentially
abundant proteins is insufficient to draw any solid conclusions. For example, if
we find that `r (nrow(sig_up))/3` of the `r nrow(sig_up)` proteins significantly
upregulated in M phase are annotated with the GO term "kinase activity", it may
seem intuitive to conclude that this biological process is important for the M-phase 
phenotype. However, if 90% of all proteins in the cell were kinases (an extreme
example), then we might expect to discover a high representation of the "kinase
activity" GO term in any protein list we end up with.

This leads us to the concept of an over-representation analysis. We wish to ask 
whether any GO terms are over-represented (i.e., present at a higher frequency 
than expected by chance) in our lists of differentially abundant proteins. In 
other words, we need to know how many proteins with a GO term *could* have
shown differential abundance in our experiment vs. how many proteins with this
GO term *did* show differential abundance in our experiment. 


<!-- ```{r} -->
<!-- ego_down <- enrichGO(gene = sig_down$protein, -->
<!--                    universe = M_G1_results$protein, -->
<!--                    OrgDb = org.Hs.eg.db, -->
<!--                    keyType = "UNIPROT", -->
<!--                    ont = "CC", -->
<!--                    pAdjustMethod = "BH", -->
<!--                    pvalueCutoff = 0.05, -->
<!--                    qvalueCutoff = 0.05, -->
<!--                    readable = TRUE) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ego_down$Description -->
<!-- ``` -->



::: {.callout-tip}
#### Key Points

- Last section of the page is a bulleted summary of the key points
:::

## References {-}